
# Loading libraries

```{r}
library(limma)
library(Seurat)         
library(SeuratObject)    
library(dplyr)
library(tidybulk)
library(tidySummarizedExperiment)
library(ggplot2)
library(ggrepel)
library(plotly)
library(tidyverse)
library(htmlwidgets)
library(DESeq2)
library(reshape2)
library(UpSetR)
library(pheatmap)
library(stringr)
library(rstatix)
library(ggsignif)
library(cbioportalR)
library(AnVIL)
library(cBioPortalData)
library(ggplot2)
library(dplyr)
library(tidyr)
library(MultiAssayExperiment)
library(msigdbr)
library(fgsea)
library(ggplot2)
library(ggvenn)
library(biomaRt)
library(org.Hs.eg.db)
suppressPackageStartupMessages({library(survival); library(survminer)})
```
# Preparing the Data

Loading the object

```{r}
# Load your Seurat object from the RDS file
seurat_obj <- readRDS("Combined.New.Object_with_Andy_cell_type_with_reduction.rds")
```

```{r}
# Specify the samples to remove
samples_to_remove <- c("Other001_D", "Other001_R")

# Find the cells that match these sample labels in the 'new_labels' metadata
cells_to_remove <- rownames(seurat_obj@meta.data)[seurat_obj$new_labels %in% samples_to_remove]

# Use the subset function to remove these cells from the Seurat object
seurat_obj <- subset(seurat_obj, cells = cells_to_remove, invert = TRUE)
```

Visualizing the Seurat Object

```{r}
# Plot UMAP colored by cluster
DimPlot(seurat_obj, reduction = "umap", group.by = "seurat_clusters")

# Plot UMAP colored by a specific metadata column, like cell type
DimPlot(seurat_obj, reduction = "umap", group.by = "dominant_cell_type")
```

```{r}
bulk <- AggregateExpression(seurat_obj, group.by = "new_labels", return.seurat = FALSE)
```

# First all D vs all R

```{r}
# The result is a list with RNA as a sparse matrix
bulk_counts <- bulk$RNA  # Access the sparse matrix with genes as rows and identity classes as columns

# Example of how to view this sparse matrix
head(bulk_counts)  
```

### Subseting the matrix

```{r}
# Convert sparse matrix to regular matrix (for DESeq2)
bulk_matrix <- as.matrix(bulk_counts)

#Filtering for low expressed genes
filtered_bulk_matrix <- bulk_matrix[apply(bulk_matrix, 1, function(x) sum(x > 10)) >= 2, ]

# Create metadata (e.g., Diagnosis and Relapse labels for your columns)
meta_data <- data.frame(
  sample = colnames(filtered_bulk_matrix),
  condition = ifelse(grepl("-D", colnames(filtered_bulk_matrix)), "Diagnosis", "Relapse"),
  stringsAsFactors = FALSE
)



```

### PCA

```{r}
# Step 1: Create the SummarizedExperiment object using filtered bulk matrix and metadata
k9.se <- SummarizedExperiment(
  assays = list(counts = as.matrix(filtered_bulk_matrix)),
  colData = meta_data
)

# Step 2: Inspect the SummarizedExperiment object
print(k9.se)

# Step 3: QC - Filter out low abundant genes and scale the abundance
# Assuming "condition" is the column we want to use as a factor of interest
counts_scaled <- k9.se %>%
  keep_abundant(factor_of_interest = condition) %>%
  scale_abundance()

# Check the scaled data
colData(counts_scaled)

# Optional: Plot the distribution of counts before and after scaling
counts_scaled %>%
  pivot_longer(cols = c("counts", "counts_scaled"), names_to = "source", values_to = "abundance") %>%
  ggplot(aes(x = abundance + 1, color = .sample)) +
  geom_density() +
  facet_wrap(~source) +
  scale_x_log10() 

# Step 4: Perform PCA
counts_scal_PCA <- counts_scaled %>%
  reduce_dimensions(method = "PCA", .dims = 3)

# Step 5: Check the PCA results
colData(counts_scal_PCA)

# Step 6: Plot 2D PCA
counts_scal_PCA %>%
  pivot_sample() %>%
  ggplot(aes(x = PC1, y = PC2, colour = condition)) +
  geom_point() +
  geom_text_repel(aes(label = .sample), show.legend = FALSE, size = 3) +
  labs(title = "PCA for all D vs all R") +
  theme_minimal() +
  theme(legend.position = "bottom")



```

### DEseq

```{r}
# Create DESeq2 dataset
dds <- DESeqDataSetFromMatrix(countData = filtered_bulk_matrix, colData = meta_data, design = ~ condition)

# Run DESeq2
dds <- DESeq(dds)

# Get DE results
res <- results(dds)
summary(res)

de_results_all_D_vs_R <- as.data.frame(res)
```
# Task 1 Compare all Diagnosis vs all Relapse

## Plotting all D vs all R

### Volcano plot

```{r}
## Plotting all D vs all R

### Volcano plot

# Define custom colors, sizes, and alphas for the plot
cols <- c("up" = "#ffad73", "down" = "#26b3ff", "ns" = "grey")
sizes <- c("up" = 2, "down" = 2, "ns" = 1)
alphas <- c("up" = 1, "down" = 1, "ns" = 0.5)

de_results_all_D_vs_R <- de_results_all_D_vs_R %>%
  rownames_to_column("GeneSymbol")


volcano_data <- de_results_all_D_vs_R %>%
  # Select only the necessary columns
  dplyr::select(GeneSymbol, log2FoldChange, padj) %>%
  # Classify genes based on fold change and adjusted p-value
  mutate(FoldChange = 2^log2FoldChange) %>%
  mutate(GeneType = case_when(
    FoldChange >= 1.5 & padj <= 0.316 ~ "up",  # Update the padj threshold to 0.316
    FoldChange <= (1/1.5) & padj <= 0.316 ~ "down",
    TRUE ~ "ns"
  ))

# Update the plot theme to ensure larger font sizes
volcano_plot <- volcano_data %>%
  ggplot(aes(x = log2(FoldChange), y = -log10(padj),
             fill = GeneType, size = GeneType, alpha = GeneType)) +
  geom_point(shape = 21, colour = "black") +
  geom_hline(yintercept = -log10(0.316), linetype = "dashed") +  # Update the y-intercept to -log10(0.316)
  geom_vline(xintercept = c(log2(0.5), log2(2)), linetype = "dashed") +
  scale_fill_manual(values = cols) +
  scale_size_manual(values = sizes) +
  scale_alpha_manual(values = alphas) +
  geom_text_repel(
    data = filter(volcano_data, padj < 0.316),  # Update the filtering for the labels
                  aes(x = log2(FoldChange), y = -log10(padj), label = GeneSymbol), 
                  size = 5,  
                  color = "darkblue", 
                  max.overlaps = Inf,
                  min.segment.length = Inf,
                  inherit.aes = FALSE) +
  labs(title = "Volcano Plot of all D vs all R with 0.316 adjpval",  # Update the title to reflect the new threshold
       x = "log2(Fold Change)", 
       y = "-log10(Adjusted P-value)") +
  theme_minimal(base_family = "Arial") +  
  theme(
    panel.background = element_rect(fill = "white", colour = NA), 
    plot.background = element_rect(fill = "white", colour = NA),  
    legend.position = "bottom",
    plot.title = element_text(size = 20, face = "bold"),   
    axis.title.x = element_text(size = 18),                
    axis.title.y = element_text(size = 18),                
    axis.text = element_text(size = 14),                   
    legend.text = element_text(size = 14),                 
    legend.title = element_text(size = 16)                 
  )

volcano_plot

# Save the plot with high resolution and proper background settings
ggsave("tasks_results/task1_allD_vs_allR/volcano_plot_allD_vs_allR_padj0.5.png", plot = volcano_plot, 
       width = 16, height = 9, dpi = 300, bg = "white")  # Ensure background is set to white

```

### violin plots for all upregulated genes

```{r}

# Step 1: Filter the significant genes based on the volcano data
significant_genes <- volcano_data %>% 
  filter(padj <= 0.316 & abs(log2(FoldChange)) >= log2(1.5)) %>%  # Include both upregulated and downregulated genes
  pull(GeneSymbol)

# Step 2: Subset the bulk matrix for these significant genes
significant_genes_data <- filtered_bulk_matrix[rownames(filtered_bulk_matrix) %in% significant_genes, ]

# Step 3: Convert the matrix to long format for ggplot
significant_genes_df <- as.data.frame(significant_genes_data) %>%
  rownames_to_column(var = "gene") %>%
  pivot_longer(cols = -gene, names_to = "sample", values_to = "expression")

# Step 4: Ensure that the meta_data has a 'subtype' column
# If it doesn't exist, create it based on the 'sample' name or other criteria
if (!"subtype" %in% colnames(meta_data)) {
  meta_data <- meta_data %>%
    mutate(subtype = case_when(
      grepl("ETV", sample) ~ "ETV",
      grepl("LH", sample) ~ "LH",
      grepl("NH", sample) ~ "NH",
      grepl("PH", sample) ~ "PH",
      TRUE ~ "Unknown"  # Assign 'Unknown' if the pattern doesn't match any subtype
    ))
}

# Step 5: Merge the expression data with the updated metadata
significant_genes_df <- merge(significant_genes_df, meta_data, by.x = "sample", by.y = "sample")

# Step 6: Label samples in "Relapse" with expression > 20
significant_genes_df <- significant_genes_df %>%
  mutate(label = ifelse(condition == "Relapse" & expression > 20, subtype, NA))  # Use subtype for labeling



```

### Saving violin plots individually

```{r}
# Define the directory to save the plots
output_dir <- "/Users/cojulian/Desktop/Tzu_projects/Matt/Integrated_D_R_curtis/tasks_results/task1_allD_vs_allR/violinplots/"

# Loop over each significant gene to create and save individual violin plots
lapply(unique(significant_genes_df$gene), function(gene_name) {
  # Subset the data for the current gene
  gene_data <- significant_genes_df %>% filter(gene == gene_name)
  
  # Create the violin plot for the current gene
  p <- ggplot(gene_data, aes(x = condition, y = expression, fill = condition)) +
    geom_violin(trim = FALSE) +
    geom_jitter(width = 0.2, size = 1.5, color = "black", alpha = 0.5) +
    geom_text_repel(aes(label = label), na.rm = TRUE, size = 5, color = "#800000") +
    labs(title = paste("Violin Plot of", gene_name, "Expression in Relapse vs Diagnosis"),
         x = "Condition",
         y = "Expression Level") +
    theme_minimal() +
    theme(legend.position = "none") +
    scale_fill_manual(values = c("Diagnosis" = "#ffad73", "Relapse" = "#26b3ff"))  # Add custom fill colors
  
  # Create the file name with format "<gene_name>_violin_plot.png"
  file_name <- paste0(gene_name, "_violin_plot.png")
  
  # Save the plot to the output directory
  ggsave(filename = file.path(output_dir, file_name), plot = p, width = 8, height = 6, dpi = 300, bg = "white")
})

```

### Calculation FPKM

```{r}
# Create a data frame with gene lengths
gene_lengths <- data.frame(
  GeneSymbol = c("MNDA", "MGST1", "LYZ", "OBSL1", "TWIST2", "PLA2G4A"),
  Length = c(18155, 262117, 5853, 21131, 62502, 160032)  # Length in base pairs
)

dds <- DESeqDataSetFromMatrix(countData = filtered_bulk_matrix, 
                              colData = meta_data, 
                              design = ~ condition)

# Step 2: Add gene lengths to the DESeqDataSet
# Assuming the 'gene_lengths' data frame contains the gene names and lengths in base pairs
mcols(dds)$basepairs <- gene_lengths$Length[match(rownames(dds), gene_lengths$GeneSymbol)]

# Step 3: Run the DESeq2 normalization
dds <- DESeq(dds)

# Step 4: Calculate FPKM values using the fpkm() function
fpkm_values_deseq2 <- fpkm(dds, robust = TRUE)  # Calculate FPKM using size factors (robust=TRUE by default)

# Step 5: Convert the matrix to a data frame for easier handling
fpkm_values_deseq2_df <- as.data.frame(fpkm_values_deseq2) %>%
  rownames_to_column(var = "GeneSymbol") %>%
  pivot_longer(cols = -GeneSymbol, names_to = "sample", values_to = "FPKM")

# Step 6: Filter the FPKM values for the genes of interest
fpkm_values_deseq2_df <- fpkm_values_deseq2_df %>%
  filter(GeneSymbol %in% c("MNDA", "MGST1", "LYZ", "OBSL1", "TWIST2", "PLA2G4A"))

# Display the head of the FPKM values
fpkm_values_deseq2_df


# Step 1: Separate the "sample" column into "sample_id" and "condition"
fpkm_values_deseq2_df <- fpkm_values_deseq2_df %>%
  separate(sample, into = c("sample_id", "condition"), sep = "-", remove = FALSE)

# Step 2: Pivot the data to get separate columns for "FPKM Diagnosis" and "FPKM Relapse"
fpkm_values_final <- fpkm_values_deseq2_df %>%
  pivot_wider(names_from = condition, values_from = FPKM, 
              names_prefix = "FPKM_") %>%
  group_by(GeneSymbol, sample_id) %>%
  summarise(FPKM_Diagnosis = sum(FPKM_D, na.rm = TRUE), 
            FPKM_Relapse = sum(FPKM_R, na.rm = TRUE), .groups = 'drop')


fpkm_values_final


#write.csv(fpkm_values_final, "fpkm_topDvsR.csv", col.names = T)
```

## Task 1.2 Subseting based on LYZ expression

```{r}
# Define the groups based on the provided table
group1_samples <- c("ETV003", "NH003", "NH006", "NH002", "NH004", "PH001", "PH002", "LH003")
group2_samples <- c("ETV001", "ETV004", "ETV005", "NH008", "NH001", "NH005", "ETV002", "PH003", "NH007", "LH002", "LH001")

# Filter meta_data for the diagnosis and relapse samples in each group
group1_meta <- meta_data %>% 
  filter(gsub("-[DR]", "", sample) %in% group1_samples)

group2_meta <- meta_data %>%
  filter(gsub("-[DR]", "", sample) %in% group2_samples)

# Subset the bulk matrix for the corresponding samples
group1_matrix <- filtered_bulk_matrix[, group1_meta$sample, drop = FALSE]
group2_matrix <- filtered_bulk_matrix[, group2_meta$sample, drop = FALSE]

# Here PCA for each

# Step 2: Run DESeq2 Analysis for Each Group Separately
# Group 1 DESeq2 analysis
dds_group1 <- DESeqDataSetFromMatrix(countData = group1_matrix, colData = group1_meta, design = ~ condition)
dds_group1 <- DESeq(dds_group1)
res_group1 <- results(dds_group1)

# Group 2 DESeq2 analysis
dds_group2 <- DESeqDataSetFromMatrix(countData = group2_matrix, colData = group2_meta, design = ~ condition)
dds_group2 <- DESeq(dds_group2)
res_group2 <- results(dds_group2)

# Step 3: Identify Significant Genes for Each Group
padj_threshold <- 0.316
log2FC_threshold <- log2(1.5)

# Significant genes for Group 1
sig_genes_group1 <- as.data.frame(res_group1) %>%
  rownames_to_column(var = "GeneSymbol") %>%
  filter(padj < padj_threshold & abs(log2FoldChange) >= log2FC_threshold)

# Significant genes for Group 2
sig_genes_group2 <- as.data.frame(res_group2) %>%
  rownames_to_column(var = "GeneSymbol") %>%
  filter(padj < padj_threshold & abs(log2FoldChange) >= log2FC_threshold)

# Step 4: Compare DE Genes Between the Groups
# Find overlapping and exclusive genes
overlap_genes <- intersect(sig_genes_group1$GeneSymbol, sig_genes_group2$GeneSymbol)
exclusive_genes_group1 <- setdiff(sig_genes_group1$GeneSymbol, sig_genes_group2$GeneSymbol)
exclusive_genes_group2 <- setdiff(sig_genes_group2$GeneSymbol, sig_genes_group1$GeneSymbol)

# Output results
list(
  overlap_genes = overlap_genes,
  exclusive_genes_group1 = exclusive_genes_group1,
  exclusive_genes_group2 = exclusive_genes_group2
)
```

### PCA group 1

```{r}

# Step 1: Create SummarizedExperiment for Group 1 using filtered matrix and metadata
group1_se <- SummarizedExperiment(
  assays = list(counts = as.matrix(group1_matrix)),
  colData = group1_meta
)

# Step 2: QC - Filter out low abundant genes and scale abundance
counts_scaled_group1 <- group1_se %>%
  keep_abundant(factor_of_interest = condition) %>%
  scale_abundance()

# Step 3: Perform PCA
counts_scaled_group1_PCA <- counts_scaled_group1 %>%
  reduce_dimensions(method = "PCA", .dims = 3)

# Step 4: Plot 2D PCA for Group 1
counts_scaled_group1_PCA %>%
  pivot_sample() %>%
  ggplot(aes(x = PC1, y = PC2, colour = condition)) +
  geom_point() +
  geom_text_repel(aes(label = .sample), show.legend = FALSE, size = 3) +
  labs(title = "PCA for Group 1 (LYZ<2FC)") +
  theme_minimal() +
  theme(legend.position = "bottom") -> PCA_group1


ggsave("tasks_results/task1.2_LYZ_grouping/PCA_Group_1_(LYZ<2FC).png", plot = PCA_group1, 
       width = 16, height = 9, dpi = 300, bg = "white")  # Ensure background is set to white

```

### Volcano plot group 1

```{r}
# Convert DESeq2 results to a data frame for Group 1
volcano_data_group1 <- as.data.frame(res_group1) %>%
  rownames_to_column("GeneSymbol") %>%
  dplyr::select(GeneSymbol, log2FoldChange, padj) %>%
  mutate(FoldChange = 2^log2FoldChange) %>%
  mutate(GeneType = case_when(
    FoldChange >= 1.5 & padj <= 0.316 ~ "up",
    FoldChange <= (1/1.5) & padj <= 0.316 ~ "down",
    TRUE ~ "ns"
  ))

# Create volcano plot for Group 1
volcano_plot_group1 <- ggplot(volcano_data_group1, aes(x = log2(FoldChange), y = -log10(padj),
                                                       fill = GeneType, size = GeneType, alpha = GeneType)) +
  geom_point(shape = 21, colour = "black") +
  geom_hline(yintercept = -log10(0.316), linetype = "dashed") +
  geom_vline(xintercept = c(log2(0.5), log2(2)), linetype = "dashed") +
  scale_fill_manual(values = cols) +
  scale_size_manual(values = sizes) +
  scale_alpha_manual(values = alphas) +
  geom_text_repel(
    data = filter(volcano_data_group1, padj < 0.316), 
    aes(x = log2(FoldChange), y = -log10(padj), label = GeneSymbol), 
    size = 5, color = "darkblue", max.overlaps = Inf, min.segment.length = Inf, inherit.aes = FALSE) +
  labs(title = "Volcano Plot for Group 1 (LYZ<2FC)", x = "log2(Fold Change)", y = "-log10(Adjusted P-value)") +
  theme_minimal(base_family = "Arial") +
  theme(
    panel.background = element_rect(fill = "white", colour = NA),
    plot.background = element_rect(fill = "white", colour = NA),
    legend.position = "bottom",
    plot.title = element_text(size = 20, face = "bold"),
    axis.title.x = element_text(size = 18),
    axis.title.y = element_text(size = 18),
    axis.text = element_text(size = 14),
    legend.text = element_text(size = 14),
    legend.title = element_text(size = 16)
  )

# Display the plot
volcano_plot_group1

ggsave("tasks_results/task1.2_LYZ_grouping/volcano_Group_1_(LYZ<2FC).png", plot = volcano_plot_group1, 
       width = 16, height = 9, dpi = 300, bg = "white")  # Ensure background is set to white

```

### PCA group 2

```{r}
# Step 1: Create SummarizedExperiment for Group 2
group2_se <- SummarizedExperiment(
  assays = list(counts = as.matrix(group2_matrix)),
  colData = group2_meta
)

# Step 2: QC - Filter out low abundant genes and scale abundance
counts_scaled_group2 <- group2_se %>%
  keep_abundant(factor_of_interest = condition) %>%
  scale_abundance()

# Step 3: Perform PCA
counts_scaled_group2_PCA <- counts_scaled_group2 %>%
  reduce_dimensions(method = "PCA", .dims = 3)

# Step 4: Plot 2D PCA for Group 2
counts_scaled_group2_PCA %>%
  pivot_sample() %>%
  ggplot(aes(x = PC1, y = PC2, colour = condition)) +
  geom_point() +
  geom_text_repel(aes(label = .sample), show.legend = FALSE, size = 3) +
  labs(title = "PCA for Group 2 (LYZ>2FC)") +
  theme_minimal() +
  theme(legend.position = "bottom") -> pca_group2

ggsave("tasks_results/task1.2_LYZ_grouping/pca_Group_2_(LYZ>2FC).png", plot = pca_group2, 
       width = 16, height = 9, dpi = 300, bg = "white")  # Ensure background is set to white

```

### Volcano plot group 2

```{r}
# Convert DESeq2 results to a data frame for Group 2
volcano_data_group2 <- as.data.frame(res_group2) %>%
  rownames_to_column("GeneSymbol") %>%
  dplyr::select(GeneSymbol, log2FoldChange, padj) %>%
  mutate(FoldChange = 2^log2FoldChange) %>%
  mutate(GeneType = case_when(
    FoldChange >= 1.5 & padj <= 0.316 ~ "up",
    FoldChange <= (1/1.5) & padj <= 0.316 ~ "down",
    TRUE ~ "ns"
  ))

# Create volcano plot for Group 2
volcano_plot_group2 <- ggplot(volcano_data_group2, aes(x = log2(FoldChange), y = -log10(padj),
                                                       fill = GeneType, size = GeneType, alpha = GeneType)) +
  geom_point(shape = 21, colour = "black") +
  geom_hline(yintercept = -log10(0.316), linetype = "dashed") +
  geom_vline(xintercept = c(log2(0.5), log2(2)), linetype = "dashed") +
  scale_fill_manual(values = cols) +
  scale_size_manual(values = sizes) +
  scale_alpha_manual(values = alphas) +
  geom_text_repel(
    data = filter(volcano_data_group2, padj < 0.316), 
    aes(x = log2(FoldChange), y = -log10(padj), label = GeneSymbol), 
    size = 5, color = "darkblue", max.overlaps = Inf, min.segment.length = Inf, inherit.aes = FALSE) +
  labs(title = "Volcano Plot for Group 2 (LYZ>2FC)", x = "log2(Fold Change)", y = "-log10(Adjusted P-value)") +
  theme_minimal(base_family = "Arial") +
  theme(
    panel.background = element_rect(fill = "white", colour = NA),
    plot.background = element_rect(fill = "white", colour = NA),
    legend.position = "bottom",
    plot.title = element_text(size = 20, face = "bold"),
    axis.title.x = element_text(size = 18),
    axis.title.y = element_text(size = 18),
    axis.text = element_text(size = 14),
    legend.text = element_text(size = 14),
    legend.title = element_text(size = 16)
  )

# Display the plot
volcano_plot_group2

ggsave("tasks_results/task1.2_LYZ_grouping/volcano_Group_2_(LYZ>2FC).png", plot = volcano_plot_group2, 
       width = 16, height = 9, dpi = 300, bg = "white")  # Ensure background is set to white


```

## Task 1.3 Avg expression of specific genes
 
### Avg expression of specific genes, zscore

```{r}
# Step 1: Select the RNA count values for GEM
gem_counts <- filtered_bulk_matrix["GEM", , drop = FALSE]  # Select the GEM row

# Step 2: Transpose to make samples as rows and genes as columns for easier manipulation
gem_counts <- t(gem_counts)

# Step 3: Convert to a data frame and add sample information
gem_counts_df <- as.data.frame(gem_counts)
gem_counts_df$Sample <- rownames(gem_counts_df)  # Add sample names as a column

# Step 4: Add metadata (Diagnosis or Relapse) for each sample
gem_counts_df <- gem_counts_df %>%
  mutate(
    Condition = ifelse(grepl("-D$", Sample), "Diagnosis", "Relapse"),  # Identify Diagnosis (D) and Relapse (R)
    Patient = sub("-[DR]$", "", Sample)  # Extract patient ID by removing the -D or -R suffix
  )

# Step 5: Combine Patient and Condition columns to format as ETV001-D, ETV001-R, etc.
gem_counts_df <- gem_counts_df %>%
  mutate(Patient_Condition = paste0(Patient, "-", substr(Condition, 1, 1)))  # Create labels like ETV001-D

# Step 6: Reorder the data frame by Patient_Condition to ensure correct order on x-axis
gem_counts_df <- gem_counts_df %>%
  arrange(Patient, Condition)

# Step 7: Plot the RNA counts of GEM with specified x-axis labels and a horizontal line at y = 0
ggplot(gem_counts_df, aes(x = Patient_Condition, y = GEM, fill = Condition)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8)) +
  geom_hline(yintercept = 0, color = "black", linetype = "dashed") +  # Add horizontal line at 0
  labs(
    title = "RNA Counts of GEM for Diagnosis and Relapse Across Patients",
    y = "RNA Count",
    x = "Sample",
    fill = "Condition"
  ) +
  scale_fill_manual(values = c("Diagnosis" = "skyblue", "Relapse" = "salmon")) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),  # Rotate x-axis labels
    axis.title.x = element_text(margin = margin(t = 10))  # Add space below x-axis labels
  )

```
### Making the loop

```{r}
# Define the target genes
target_genes <- c("GEM", "ZNF692", "USE1", "ZNF22", "INHBE", "CDK4", "RASL11A",
                  "CYP7B1", "SLC35F1", "NFKBIE", "ZBTB10", "CTSB", "CNBD2",
                  "CDC42EP4", "OLAH", "FAM76B")

# Function to create a plot for a single gene
plot_gene_counts <- function(gene) {
  # Step 1: Select the RNA count values for the gene
  gene_counts <- filtered_bulk_matrix[gene, , drop = FALSE]  # Select the gene row
  
  # Step 2: Transpose to make samples as rows and genes as columns for easier manipulation
  gene_counts <- t(gene_counts)
  
  # Step 3: Convert to a data frame and add sample information
  gene_counts_df <- as.data.frame(gene_counts)
  gene_counts_df$Sample <- rownames(gene_counts_df)  # Add sample names as a column
  
  # Step 4: Add metadata (Diagnosis or Relapse) for each sample
  gene_counts_df <- gene_counts_df %>%
    mutate(
      Condition = ifelse(grepl("-D$", Sample), "Diagnosis", "Relapse"),  # Identify Diagnosis (D) and Relapse (R)
      Patient = sub("-[DR]$", "", Sample)  # Extract patient ID by removing the -D or -R suffix
    )
  
  # Step 5: Combine Patient and Condition columns to format as ETV001-D, ETV001-R, etc.
  gene_counts_df <- gene_counts_df %>%
    mutate(Patient_Condition = paste0(Patient, "-", substr(Condition, 1, 1)))  # Create labels like ETV001-D
  
  # Step 6: Reorder the data frame by Patient_Condition to ensure correct order on x-axis
  gene_counts_df <- gene_counts_df %>%
    arrange(Patient, Condition)
  
  # Step 7: Plot the RNA counts for the gene with specified x-axis labels and a horizontal line at y = 0
  ggplot(gene_counts_df, aes(x = Patient_Condition, y = get(gene), fill = Condition)) +
    geom_bar(stat = "identity", position = position_dodge(width = 0.8)) +
    geom_hline(yintercept = 0, color = "black", linetype = "dashed") +  # Add horizontal line at 0
    labs(
      title = paste("RNA Counts of", gene, "for Diagnosis and Relapse Across Patients"),
      y = "RNA Count",
      x = "Sample",
      fill = "Condition"
    ) +
    scale_fill_manual(values = c("Diagnosis" = "skyblue", "Relapse" = "salmon")) +
    theme_minimal() +
    theme(
       plot.background = element_rect(fill = "white", color = NA),  # White plot background
      panel.background = element_rect(fill = "white", color = NA),  # White panel background
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),  # Rotate x-axis labels
      axis.title.x = element_text(margin = margin(t = 10))  # Add space below x-axis labels
    )
}

# Apply the function to each gene in target_genes and save the plots in a list
plots_list <- map(target_genes, plot_gene_counts)

# Optional: Print or save each plot
# Example to display a plot, e.g., the first plot
print(plots_list[[1]])

# Specify the output directory for the plots
output_dir <- "/Users/cojulian/Desktop/Tzu_projects/Matt/Integrated_D_R_curtis/tasks_results/task1.3_avg_expression_rnacounts"

# Ensure the directory exists
if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}

# Save each plot as a PNG in the specified output directory
for (i in seq_along(target_genes)) {
  ggsave(
    filename = file.path(output_dir, paste0("plot_", target_genes[i], ".png")),
    plot = plots_list[[i]],
    width = 10,
    height = 6,
    dpi = 300  # High resolution for PNG
  )
}

# This has been a test
```

# Filtering bulk_matrix by stjudes_filtered

```{r}
# Extract the gene names from lowexp_filt_heatmap_data
genes_to_keep <- rownames(lowexp_filt_heatmap_data)

# Filter filtered_bulk_matrix to retain only the genes in genes_to_keep
filtered_bulk_matrix_subset <- filtered_bulk_matrix[rownames(filtered_bulk_matrix) %in% genes_to_keep, ]

# Check the result
dim(filtered_bulk_matrix_subset) 

# Calculate Z-scores for each gene (row-wise)
filtered_bulk_matrix_subset_zscore_matrix <- t(apply(filtered_bulk_matrix_subset, 1, function(x) {
  (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)
}))

```

```{r}
library(pheatmap)

# Generate the heatmap
library(pheatmap)

library(pheatmap)

# Generate the heatmap
pheatmap::pheatmap(
  filtered_bulk_matrix_subset_zscore_matrix,
  cluster_rows = TRUE,       # Cluster rows (genes)
  cluster_cols = TRUE,       # Cluster columns (samples)
  main = "Z-Scores of Genes (Filtered for Low Expression)",  # Heatmap title
  show_rownames = TRUE,       # Show gene names
  show_colnames = TRUE,       # Show sample names
  color = colorRampPalette(c("blue", "white", "red"))(50),  # Color gradient
  fontsize_col = 8,           # Font size for column names
  fontsize_row = 6            # Font size for row names
)
```

## Rownormalizing this

```{r}
# Step 1: Calculate Z-scores for all genes in filtered_bulk_matrix
filtered_bulk_matrix_zscore <- t(apply(filtered_bulk_matrix, 1, function(x) {
  (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)
}))

# Step 2: Filter for genes of interest
# Extract the gene names from lowexp_filt_heatmap_data
genes_to_keep <- rownames(lowexp_filt_heatmap_data)

# Subset the Z-score matrix to retain only the genes in genes_to_keep
filtered_bulk_matrix_subset_zscore <- filtered_bulk_matrix_zscore[rownames(filtered_bulk_matrix_zscore) %in% genes_to_keep, ]

# Check the result
dim(filtered_bulk_matrix_subset_zscore)

library(pheatmap)

# Set 'Gene' as row names and convert to a numeric matrix
filtered_bulk_heatmap_data <- filtered_bulk_matrix_subset_zscore %>%
  as.matrix()

# Generate the heatmap
pheatmap::pheatmap(
  filtered_bulk_heatmap_data,
  cluster_rows = T,       # Cluster rows (genes)
  cluster_cols = T,       # Cluster columns (samples)
  main = "Z-Score Heatmap of Filtered Bulk Matrix",  # Heatmap title
  show_rownames = TRUE,       # Show gene names (set to FALSE if too many genes)
  show_colnames = TRUE,       # Show sample names
  color = colorRampPalette(c("blue", "white", "red"))(50),  # Color gradient
  fontsize_col = 8,           # Font size for column names
  fontsize_row = 6            # Font size for row names
)

```


# Task 2 D vs R per subtype

Make a volcano plot per subtype, get the files and make a venn diagram or upset of commons

## PCA

```{r}
# List of subtypes
subtypes <- unique(meta_data$subtype)

# Create a list to store PCA results for each subtype
pca_results_list <- list()

# Loop through each subtype
for (subtype in subtypes) {
  # Step 1: Filter the bulk matrix and metadata for the current subtype
  subtype_meta_data <- meta_data[meta_data$subtype == subtype, ]
  subtype_bulk_matrix <- filtered_bulk_matrix[, subtype_meta_data$sample]
  
  # Step 2: Create the SummarizedExperiment object for the current subtype
  subtype_se <- SummarizedExperiment(
    assays = list(counts = as.matrix(subtype_bulk_matrix)),
    colData = subtype_meta_data
  )
  
  # Step 3: QC - Filter out low abundant genes and scale the abundance
  counts_scaled <- subtype_se %>%
    keep_abundant(factor_of_interest = condition) %>%  # Use {{}} for symbolic reference
    scale_abundance()
  
  # Step 4: Perform PCA
  counts_scal_PCA <- counts_scaled %>%
    reduce_dimensions(method = "PCA", .dims = 3)
  
  # Step 5: Store the PCA results
  pca_results_list[[subtype]] <- counts_scal_PCA
  
  # Step 6: Plot the PCA results for the current subtype
  pca_plot <- counts_scal_PCA %>%
    pivot_sample() %>%
    ggplot(aes(x = PC1, y = PC2, colour = condition)) +
    geom_point() +
    geom_text_repel(aes(label = .sample), show.legend = FALSE, size = 3) +
    labs(title = paste0("PCA for D vs R per Subtype: ", subtype),
         x = "PC1", y = "PC2") +
    theme_minimal() +
    theme(legend.position = "bottom")
  
  # Step 7: Print the plot (or save it)
  print(pca_plot)
}

```

Remove Other

## Function for running dds

```{r}
# Initialize an empty list to store DE results per subtype
de_results_subtype_list <- list()

# Loop over each subtype and perform DE analysis for Diagnosis vs Relapse
for (subtype in unique(meta_data$subtype)) {
  
  # Subset the metadata for the current subtype
  subtype_meta_data <- meta_data[meta_data$subtype == subtype, ]
  
  # Subset the bulk matrix for the current subtype based on samples in metadata
  bulk_subset <- filtered_bulk_matrix[, subtype_meta_data$sample]
  
  # Create a DESeq2 dataset
  dds <- DESeqDataSetFromMatrix(
    countData = bulk_subset,
    colData = subtype_meta_data,
    design = ~ condition
  )
  
  # Run DESeq2 analysis
  dds <- DESeq(dds)
  
  # Extract results, order by adjusted p-value
  de_results <- results(dds, alpha = 0.05) %>%
    as.data.frame() %>%
    arrange(padj) 
  
  # Store the DE results in the list
  de_results_subtype_list[[subtype]] <- de_results
}

# Example: Check DE results for a specific subtype (e.g., ETV)
de_results_subtype_list$ETV
```

### Volcano Plot Task 2

```{r, fig.align='default'}
# Update the color, size, and alpha settings for the volcano plot
cols <- c("up" = "#ffad73", "down" = "#26b3ff", "ns" = "grey")
sizes <- c("up" = 2, "down" = 2, "ns" = 1)
alphas <- c("up" = 1, "down" = 1, "ns" = 0.5)

# Function to create a volcano plot with an adjusted p-value cutoff of 0.316
create_volcano_plot <- function(de_results, subtype) {
  # Prepare data for the volcano plot
  volcano_data <- de_results %>%
    rownames_to_column(var = "GeneSymbol") %>%
    drop_na() %>%
    dplyr::select(GeneSymbol, log2FoldChange, padj) %>%
    mutate(FoldChange = 2^log2FoldChange) %>%
    mutate(GeneType = case_when(
      FoldChange >= 1.5 & padj <= 0.316 ~ "up",  # Update the padj threshold to 0.316
      FoldChange <= (1/1.5) & padj <= 0.316 ~ "down",
      TRUE ~ "ns"
    ))

  # Create the volcano plot
  p <- ggplot(volcano_data, aes(x = log2(FoldChange), y = -log10(padj), 
                                fill = GeneType, 
                                size = GeneType, 
                                alpha = GeneType)) +
    geom_point(shape = 21, colour = "black") +
    geom_hline(yintercept = -log10(0.316), linetype = "dashed") +  # Update the y-intercept to -log10(0.316)
    geom_vline(xintercept = c(log2(1/1.5), log2(1.5)), linetype = "dashed") +
    scale_fill_manual(values = cols) +
    scale_size_manual(values = sizes) +
    scale_alpha_manual(values = alphas) +
    geom_text_repel(data = filter(volcano_data, padj < 0.316 & abs(log2(FoldChange)) >= log2(1.5)), # Update the filtering for the labels
                    aes(x = log2(FoldChange), y = -log10(padj), label = GeneSymbol),
                    size = 5,  # Increase label size
                    color = "darkblue",
                    max.overlaps = 10,  # Limit the number of label overlaps
                    box.padding = 0.5,  # Increase padding to reduce clutter
                    min.segment.length = 0.5,
                    inherit.aes = FALSE) +
    labs(title = paste("Volcano Plot of D vs R for", subtype, "with 0.316 adjpval"),
         x = "Log2 Fold Change",
         y = "-Log10 Adjusted P-Value") +
    theme_minimal(base_family = "Arial") +
    theme(
      panel.backdground = element_rect(fill = "white", colour = NA),
      plot.background = element_rect(fill = "white", colour = NA),
      legend.position = "bottom",
      plot.title = element_text(size = 20, face = "bold"),
      axis.title.x = element_text(size = 18),
      axis.title.y = element_text(size = 18),
      axis.text = element_text(size = 14),
      legend.text = element_text(size = 14),
      legend.title = element_text(size = 16)
    )

  return(p)
}

# Generate and store volcano plots for each subtype in de_results_subtype_list
volcano_plots <- lapply(names(de_results_subtype_list), function(subtype) {
  de_results <- de_results_subtype_list[[subtype]]
  create_volcano_plot(de_results, subtype)
})

# Optionally, set names to the list of plots
names(volcano_plots) <- names(de_results_subtype_list)

volcano_plots

```

### Saving D vs R persubtype

```{r}
# Define the directory to save the Task 2 volcano plots
output_dir <- "/Users/cojulian/Desktop/Tzu_projects/Matt/Integrated_D_R_curtis/tasks_results/task2_DvsR_per_subtype"

# Save each volcano plot with a specified filename
lapply(names(volcano_plots), function(subtype) {
  # Get the plot for the given subtype
  plot <- volcano_plots[[subtype]]
  
  # Create the file name with format "D_vs_R_<subtype>_0.5padj.png"
  file_name <- paste0("D_vs_R_", subtype, "_0.316padj.png") 
  
  # Save the plot using ggsave with full HD settings and white background
  ggsave(filename = file.path(output_dir, file_name), 
         plot = plot, 
         width = 16, height = 9, dpi = 300, bg = "white")
})

```

### Trying upset plot

```{r}
# Function to identify significant genes based on updated thresholds
get_significant_genes <- function(de_results, padj_threshold = 0.316, log2FC_threshold = 1.5) {
  # Filter the significant genes and get their row names
  significant_genes <- de_results %>%
    filter(padj < padj_threshold & abs(log2FoldChange) >= log2FC_threshold) %>%
    rownames()  # Get the row names as the gene names
  return(significant_genes)
}

# Create a list of significant genes for each subtype using the updated thresholds
significant_genes_list <- lapply(de_results_subtype_list, get_significant_genes)

# Prepare the data for the UpSet plot
all_genes <- unique(unlist(significant_genes_list))
upset_data <- data.frame(
  Gene = all_genes,
  ETV = as.integer(all_genes %in% significant_genes_list$ETV),
  LH = as.integer(all_genes %in% significant_genes_list$LH),
  NH = as.integer(all_genes %in% significant_genes_list$NH),
  PH = as.integer(all_genes %in% significant_genes_list$PH)
)

# Remove the Gene column for plotting
upset_data <- upset_data %>%dplyr::select(-Gene)

# Define the directory to save the Task 2 UpSet plot
output_dir <- "/Users/cojulian/Desktop/Tzu_projects/Matt/Integrated_D_R_curtis/tasks_results/task2_DvsR_per_subtype"

# Save the UpSet plot with the updated thresholds in the filename
output_file <- file.path(output_dir, "UpSetPlot_DvsR_0.316padj_1.5FC.png")

# Create and save the UpSet plot
png(filename = output_file, width = 16, height = 9, units = "in", res = 300)
upset(upset_data, sets = c("ETV", "LH", "NH", "PH"), order.by = "freq", 
      mainbar.y.label = "Intersection Size", sets.x.label = "Set Size")
dev.off()


```

```{r, fig.height= 10}
# Step 1: Extract the top 20 significant genes from the ETV results, ordered by adjusted p-value
top20_sig_genes <- de_results_subtype_list$ETV %>%
  arrange(padj) %>%
  head(n = 20) %>%
  rownames()

# Step 2: Extract expression values for these genes from the filtered pseudobulk matrix
# Subset the matrix to include only the top 20 significant genes
top20_sig_counts <- filtered_bulk_matrix[rownames(filtered_bulk_matrix) %in% top20_sig_genes, ]

# Step 3: Convert the matrix to long format for ggplot2
top20_sig_df <- as.data.frame(top20_sig_counts)
top20_sig_df$gene <- rownames(top20_sig_df)

# Melt the data into long format using reshape2::melt
top20_sig_df <- melt(top20_sig_df, id.vars = "gene", variable.name = "sample", value.name = "expression")

# Step 4: Merge with metadata to get 'condition' (Diagnosis/Relapse) and subtype information
top20_sig_df <- merge(top20_sig_df, meta_data, by.x = "sample", by.y = "sample")

# Step 5: Generate the violin plot with ggplot2
ggplot(top20_sig_df, aes(x = condition, y = expression, fill = condition)) +
  geom_violin(trim = FALSE) +  # Draw the violin plots
  geom_jitter(shape = 16, position = position_jitter(0.2), size = 1, color = "black", alpha = 0.6) +  # Add individual points
  facet_wrap(~ gene, scales = "free_y") +  # Separate each gene into its own panel
  scale_y_continuous(trans = 'log10') +  # Use log scale for the y-axis
  labs(title = "Top 20 Significant DE Genes for ETV",
       y = "log10 of expression level",
       x = "Condition (Diagnosis vs Relapse)") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 20),  # Increase title text size
    axis.text.x = element_text(angle = 45, hjust = 1, size = 12),  # Increase x-axis text size
    axis.text.y = element_text(size = 12),  # Increase y-axis text size
    axis.title = element_text(size = 14),  # Increase axis titles size
    strip.background = element_rect(fill = "grey80", color = NA),  # Set grey background for gene names
    strip.text = element_text(color = "black", face = "bold", size = 14)  # Increase gene labels text size
  ) +
  scale_fill_manual(values = c("Diagnosis" = "#ffad73", "Relapse" = "#26b3ff"))  # Customize color scheme

```

##### Here make a volcano plot with a fc of 1.5

#### D vs D by Subtype

One pca only of Ds and also FindMarkers

```{r}
# Add subtype column to meta_data based on sample names
meta_data$subtype <- ifelse(grepl("PH", meta_data$sample), "PH", 
                            ifelse(grepl("ETV", meta_data$sample), "ETV",
                                   ifelse(grepl("LH", meta_data$sample), "LH", "NH")))

# Filter the metadata to only include Diagnosis samples
diagnosis_meta_data <- meta_data[meta_data$condition == "Diagnosis", ]
```

```{r}
# List of subtypes (for Diagnosis samples only)
diagnosis_subtypes <- unique(diagnosis_meta_data$subtype)
# Filter the bulk matrix to only include Diagnosis samples
diagnosis_bulk_matrix <- filtered_bulk_matrix[, diagnosis_meta_data$sample]

# Step 1: Create the SummarizedExperiment object for all Diagnosis samples
diagnosis_se <- SummarizedExperiment(
  assays = list(counts = as.matrix(diagnosis_bulk_matrix)),
  colData = diagnosis_meta_data
)

# Step 2: QC - Filter out low abundant genes and scale the abundance
counts_scaled <- diagnosis_se %>%
  keep_abundant(factor_of_interest = condition) %>%
  scale_abundance()

# Step 3: Perform PCA on the combined Diagnosis samples
counts_scal_PCA <- counts_scaled %>%
  reduce_dimensions(method = "PCA", .dims = 3)

# Step 4: Plot the PCA results, color-coding by subtype
pca_plot <- counts_scal_PCA %>%
  pivot_sample() %>%
  ggplot(aes(x = PC1, y = PC2, colour = subtype)) +
  geom_point() +
  geom_text_repel(aes(label = .sample), show.legend = FALSE, size = 3) +
  labs(title = "PCA for All Diagnosis Samples",
       x = "PC1", y = "PC2") +
  theme_minimal() +
  theme(legend.position = "bottom")

# Step 5: Print the PCA plot
print(pca_plot)

```

```{r}
# Initialize an empty list to store DE results for Diagnosis comparisons between subtypes
de_results_subtype_diagnosis_list <- list()

# Define pairs of subtypes to compare, e.g., ETV vs LH Diagnosis
subtype_comparison_pairs <- list(c("ETV", "LH"), c("ETV", "NH"), c("LH", "NH"), c("PH", "ETV"), c("PH", "LH"))

# Loop through each pair of subtypes to perform DE analysis for Diagnosis samples
for (pair in subtype_comparison_pairs) {
  subtype1 <- pair[1]
  subtype2 <- pair[2]
  
  # Subset the metadata for the two subtypes
  subtype1_meta_data <- diagnosis_meta_data[diagnosis_meta_data$subtype == subtype1, ]
  subtype2_meta_data <- diagnosis_meta_data[diagnosis_meta_data$subtype == subtype2, ]
  
  # Combine the metadata for the two subtypes
  combined_meta_data <- rbind(subtype1_meta_data, subtype2_meta_data)
  
  # Subset the bulk matrix for the two subtypes
  combined_bulk_matrix <- filtered_bulk_matrix[, combined_meta_data$sample]
  
  # Create a DESeq2 dataset
  dds <- DESeqDataSetFromMatrix(
    countData = combined_bulk_matrix,
    colData = combined_meta_data,
    design = ~ subtype  # Compare subtypes instead of condition
  )
  
  # Run DESeq2 analysis
  dds <- DESeq(dds)
  
  # Extract results, order by adjusted p-value
  de_results <- results(dds, contrast = c("subtype", subtype1, subtype2), alpha = 0.05) %>%
    as.data.frame() %>%
    arrange(padj)
  
  # Store the DE results in the list
  comparison_name <- paste0(subtype1, "_vs_", subtype2)
  de_results_subtype_diagnosis_list[[comparison_name]] <- de_results
}

# Example: Check DE results for a specific comparison (e.g., ETV vs LH)
head(de_results_subtype_diagnosis_list$ETV_vs_LH)

```

## PERSONAL TRYING

```{r}
lh_meta_data <- meta_data[meta_data$subtype == "LH", ]
lh_bulk_matrix <- filtered_bulk_matrix[, lh_meta_data$sample]
```

```{r}
lh_dds <- DESeqDataSetFromMatrix(
  countData = lh_bulk_matrix,
  colData = lh_meta_data,
  design = ~ condition
)
lh_dds <- DESeq(lh_dds)
lh_de_results <- results(lh_dds, alpha = 0.05) %>%
  as.data.frame() %>%
  arrange(padj)

```
```{r}
downregulated_genes <- lh_de_results %>%
  filter(log2FoldChange < 0 & padj < 0.316) %>%
  rownames()
```
```{r, fig.height=10, fig.width=10}
library(clusterProfiler)
go_results <- enrichGO(gene = downregulated_genes,
                       OrgDb = org.Hs.eg.db,
                       keyType = "SYMBOL",
                       ont = "BP", # Biological Process
                       pvalueCutoff = 0.316)
barplot(go_results, showCategory = 10, title = "Top GO Terms for Downregulated Genes")

```

```{r}
# Assuming you have a data frame `lh_de_results` with gene expression results
# Filter and rank genes
ranked_genes <- lh_de_results %>%
  rownames_to_column("Gene") %>% 
  dplyr::select(Gene, log2FoldChange) %>%
  arrange(desc(log2FoldChange))

# Create a named vector of ranked genes
gene_ranks <- ranked_genes %>%
  deframe()

# Break ties in ranked genes by adding small random noise
set.seed(123)  # For reproducibility
gene_ranks <- gene_ranks + rnorm(length(gene_ranks), mean = 0, sd = 1e-5)
```

```{r}

# Retrieve Hallmark gene sets for Homo sapiens
hallmark_pathways <- msigdbr(species = "Homo sapiens", category = "H")

# Format pathways for fgsea
hallmark_pathway_list <- hallmark_pathways %>%
  split(x = .$gene_symbol, f = .$gs_name)
```

```{r}
# Remove non-finite values
gene_ranks <- gene_ranks[is.finite(gene_ranks)]

# Alternatively, replace non-finite values with zero
# gene_ranks[!is.finite(gene_ranks)] <- 0

# Run fgseaMultilevel for better precision
fgsea_results <- fgsea(
  pathways = hallmark_pathway_list,
  stats = gene_ranks,
  minSize = 15,
  maxSize = 500,
  eps = 0  # Increases precision for very small p-values
)
```

```{r}
# Convert results to a tidy tibble and filter significant pathways
fgsea_tidy <- fgsea_results %>%
  as_tibble() %>%
  arrange(padj)

significant_pathways <- fgsea_tidy %>%
  filter(padj < 0.316)
```
```{r}
# Bar plot of top 20 significant pathways by NES
fgsea_tidy %>%
  filter(NES < 0) %>%          # Filter significant pathways
  slice_min(padj, n = 20) %>%     # Select top 20 pathways by adjusted p-value
  ggplot(aes(x = reorder(pathway, NES), y = NES, fill = NES > 0)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top Enriched Hallmark Pathways",
    x = "Pathway",
    y = "Normalized Enrichment Score (NES)"
  ) +
  scale_fill_manual(values = c("TRUE" = "#61BB6F", "FALSE" = "#EC7012")) +
  theme_minimal()
```

## Figure 2 producing

```{r}

```



# Task 3 Volcano plot D within each subtype

```{r}
# Define colors, sizes, and alphas for the volcano plot
cols <- c("up" = "#ffad73", "down" = "#26b3ff", "ns" = "grey")
sizes <- c("up" = 2, "down" = 2, "ns" = 1)
alphas <- c("up" = 1, "down" = 1, "ns" = 0.5)
```

```{r}
# Function to create a volcano plot with updated thresholds for Task 3
create_volcano_plot_diagnosis <- function(de_results, comparison_name) {
  # Prepare data for the volcano plot
  volcano_data <- de_results %>%
    rownames_to_column(var = "GeneSymbol") %>% 
    drop_na() %>%
    mutate(FoldChange = 2^log2FoldChange) %>%
    mutate(GeneType = case_when(
      FoldChange >= 1.5 & padj <= 0.316 ~ "up",  # Update the padj threshold to 0.316
      FoldChange <= (1/1.5) & padj <= 0.316 ~ "down",
      TRUE ~ "ns"
    ))

  # Create the volcano plot
  p <- ggplot(volcano_data, aes(x = log2(FoldChange), y = -log10(padj), 
                                fill = GeneType, 
                                size = GeneType, 
                                alpha = GeneType)) +
    geom_point(shape = 21, colour = "black") +
    geom_hline(yintercept = -log10(0.316), linetype = "dashed") +  # Update the y-intercept to -log10(0.316)
    geom_vline(xintercept = c(log2(1/1.5), log2(1.5)), linetype = "dashed") +
    scale_fill_manual(values = cols) +
    scale_size_manual(values = sizes) +
    scale_alpha_manual(values = alphas) +
    geom_text_repel(
      data = filter(volcano_data, padj < 0.316 & abs(log2(FoldChange)) >= log2(1.5)),  # Update the filtering for the labels
      aes(x = log2(FoldChange), y = -log10(padj), label = GeneSymbol),
      size = 5,  
      color = "darkblue",
      max.overlaps = 0.5,
      box.padding = 0.5,
      min.segment.length = Inf,
      inherit.aes = FALSE
    ) +
    labs(title = paste("Volcano Plot for", comparison_name, "with 0.316 adjpval"),
         x = "Log2 Fold Change",
         y = "-Log10 Adjusted P-Value") +
    theme_minimal(base_family = "Arial") +
    theme(
      panel.background = element_rect(fill = "white", colour = NA),
      plot.background = element_rect(fill = "white", colour = NA),
      legend.position = "bottom",
      plot.title = element_text(size = 20, face = "bold"),
      axis.title.x = element_text(size = 18),
      axis.title.y = element_text(size = 18),
      axis.text = element_text(size = 14),
      legend.text = element_text(size = 14),
      legend.title = element_text(size = 16)
    )

  return(p)
}

# Generate and store volcano plots for each subtype comparison in de_results_subtype_diagnosis_list
volcano_plots_diagnosis <- lapply(names(de_results_subtype_diagnosis_list), function(comparison_name) {
  de_results <- de_results_subtype_diagnosis_list[[comparison_name]]
  create_volcano_plot_diagnosis(de_results, comparison_name)
})

# Optionally, set names to the list of plots
names(volcano_plots_diagnosis) <- names(de_results_subtype_diagnosis_list)

volcano_plots_diagnosis


```

Saving them

```{r}
# Define the directory to save the plots
output_dir <- "/Users/cojulian/Desktop/Tzu_projects/Matt/Integrated_D_R_curtis/tasks_results/task3_all_Ds_per_subtpye"

# Save each volcano plot with a specified filename
lapply(names(volcano_plots_diagnosis), function(comparison_name) {
  # Get the plot
  plot <- volcano_plots_diagnosis[[comparison_name]]
  
  # Create the file name with format "D_<comparison_name>_0.5padj.png"
  file_name <- paste0("D_", comparison_name, "_0.316padj.png")
  
  # Save the plot using ggsave with full HD settings and white background
  ggsave(filename = file.path(output_dir, file_name), 
         plot = plot, 
         width = 16, height = 9, dpi = 300, bg = "white")
})
```

## Trying the heatmap

```{r}
# Step 1: Filter the bulk matrix to keep only diagnosis samples
diagnosis_samples <- meta_data %>%
  filter(condition == "Diagnosis") %>%
  pull(sample)

diagnosis_matrix <- filtered_bulk_matrix[, diagnosis_samples, drop = FALSE]

# Step 2: Create an empty list to store the columns for each subtype with averages
extended_matrix <- as.data.frame(diagnosis_matrix)  # Start with the filtered matrix

# 1. Filter for ETV subtype columns
etv_matrix <- extended_matrix %>%dplyr::select(starts_with("ETV"))

# 2. Filter for LH subtype columns
lh_matrix <- extended_matrix %>%dplyr::select(starts_with("LH"))

# 3. Filter for NH subtype columns
nh_matrix <- extended_matrix %>%dplyr::select(starts_with("NH"))

# 4. Filter for PH subtype columns
ph_matrix <- extended_matrix %>%dplyr::select(starts_with("PH"))

# Display structure to confirm
str(etv_matrix)
str(lh_matrix)
str(nh_matrix)
str(ph_matrix)
```
## Trying only for ETV

```{r}
# List of matrices and their respective average column names
matrices <- list(ETV = etv_matrix, LH = lh_matrix, NH = nh_matrix, PH = ph_matrix)
average_col_names <- c("ETV_avg", "LH_avg", "NH_avg", "PH_avg")

# Function to add average column to each matrix
add_avg_column <- function(matrix, avg_col_name, prefix) {
  matrix %>%
    dplyr::mutate(!!avg_col_name := rowMeans(dplyr::select(., starts_with(prefix)), na.rm = TRUE))
}

# Apply the function and assign each output to a separate variable
etv_matrix <- add_avg_column(etv_matrix, "ETV_avg", "ETV")
lh_matrix <- add_avg_column(lh_matrix, "LH_avg", "LH")
nh_matrix <- add_avg_column(nh_matrix, "NH_avg", "NH")
ph_matrix <- add_avg_column(ph_matrix, "PH_avg", "PH")

# Check structures to confirm each data frame has the new column
str(etv_matrix)
str(lh_matrix)
str(nh_matrix)
str(ph_matrix)
```

## Merging

```{r}
# Extract only the average columns from each matrix and rename them to subtype names
etv_avg <- etv_matrix %>% dplyr::select(ETV_avg)
lh_avg <- lh_matrix %>% dplyr::select(LH_avg)
nh_avg <- nh_matrix %>% dplyr::select(NH_avg)
ph_avg <- ph_matrix %>% dplyr::select(PH_avg)

# Combine the average columns into a single data frame, keeping row names intact
combined_avg_expression <- data.frame(
  ETV = etv_avg$ETV_avg,
  LH = lh_avg$LH_avg,
  NH = nh_avg$NH_avg,
  PH = ph_avg$PH_avg,
  row.names = rownames(etv_matrix)
)

# Check the structure to confirm the combined format
str(combined_avg_expression)

```



```{r, fig.height=25}
# Z-score normalization of each gene across the subtypes
zscore_matrix <- t(apply(combined_avg_expression, 1, function(x) (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)))

# Calculate the variance of each gene across subtypes
gene_variances <- apply(zscore_matrix, 1, var, na.rm = TRUE)

# Get the names of the top 100 most variable genes
top_100_genes <- names(sort(gene_variances, decreasing = TRUE))[1:500]

# Subset the z-score matrix to include only the top 100 genes
top_100_zscore_matrix <- zscore_matrix[top_100_genes, ]

# Plot the heatmap with the top 100 most variable genes
pheatmap(
  top_100_zscore_matrix,
  cluster_rows = T,            # Cluster genes by their z-score patterns
  cluster_cols = F,            # Cluster subtypes
  main = "Top 50 Most Variable Genes (Z-Score Normalized) Across Subtypes",
  show_rownames = F,           # Show gene names for clarity
  show_colnames = TRUE,           # Show subtype labels
  color = colorRampPalette(c("blue", "white", "red"))(50), # Blue = low, Red = high
  fontsize_col = 35,
  fontsize_row = 35,
  fontsize = 35
)
```



# Producing figure 1 part 1

```{r}
# Z-score normalization of each gene across the subtypes
zscore_matrix <- t(apply(combined_avg_expression, 1, function(x) (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)))

# Calculate the variance of each gene across subtypes
gene_variances <- apply(zscore_matrix, 1, var, na.rm = TRUE)

# Get the names of the top 1000 most variable genes
top_1000_genes <- names(sort(gene_variances, decreasing = TRUE))[1:1000]

# Subset the z-score matrix to include only the top 1000 genes
top_1000_zscore_matrix <- zscore_matrix[top_1000_genes, ]

# Rank by LH and NH Z-scores
ranked_by_LH <- top_1000_zscore_matrix[order(-top_1000_zscore_matrix[, "LH"]), ]
ranked_by_NH <- top_1000_zscore_matrix[order(-top_1000_zscore_matrix[, "NH"]), ]


# Create the directory if it doesn't exist
output_dir <- "/Users/cojulian/Desktop/Tzu_projects/Matt/Integrated_D_R_curtis/tasks_results/figure_1"
if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}

# Function to save pheatmap as SVG
save_pheatmap_svg <- function(pheatmap_obj, filename) {
  svg(filename)
  grid::grid.draw(pheatmap_obj$gtable)
  dev.off()
}

# Plot heatmap ranked by LH and save as SVG
p1 <- pheatmap::pheatmap(
  ranked_by_LH,
  cluster_rows = FALSE,  # Do not cluster rows to preserve ranking order
  cluster_cols = FALSE,  # Do not cluster columns
  main = "Top 1000 Variable Genes at Diagnosis Ranked by LH Z-Score (Single-Cell)",
  show_rownames = FALSE, # Hide gene names for clarity
  show_colnames = TRUE,  # Show subtype labels
  color = colorRampPalette(c("blue", "white", "red"))(50), # Color scheme
  fontsize_col = 10,
  fontsize_row = 10
)
save_pheatmap_svg(p1, file.path(output_dir, "sc_top1000_LHranked.svg"))

# Plot heatmap ranked by NH and save as SVG
p2 <- pheatmap::pheatmap(
  ranked_by_NH,
  cluster_rows = FALSE,  # Do not cluster rows to preserve ranking order
  cluster_cols = FALSE,  # Do not cluster columns
  main = "Top 1000 Variable Genes at Diagnosis Ranked by NH Z-Score (Single-Cell)",
  show_rownames = FALSE, # Hide gene names for clarity
  show_colnames = TRUE,  # Show subtype labels
  color = colorRampPalette(c("blue", "white", "red"))(50), # Color scheme
  fontsize_col = 10,
  fontsize_row = 10
)
save_pheatmap_svg(p2, file.path(output_dir, "sc_top1000_NHranked.svg"))




```

### Loking for KIF9

```{r}
# Z-score normalization of each gene across the subtypes
zscore_matrix <- t(apply(combined_avg_expression, 1, function(x) (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)))

# Calculate the variance of each gene across subtypes
gene_variances <- apply(zscore_matrix, 1, var, na.rm = TRUE)

# Get the names of the top 500 most variable genes
top_500_genes <- names(sort(gene_variances, decreasing = TRUE))[1:500]

# Subset the z-score matrix to include only the top 500 genes
top_500_zscore_matrix <- zscore_matrix[top_500_genes, ]

# Filter the matrix to include only the row for gene "KIF9"
kif9_data <- zscore_matrix["KIF9", , drop = FALSE]

# Plot the heatmap for the gene "KIF9"
pheatmap::pheatmap(
  kif9_data,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  main = "Heatmap for gene KIF9 on Diagnosis single cell",
  show_rownames = TRUE,
  show_colnames = TRUE,
  color = colorRampPalette(c("blue", "white", "red"))(50),
  fontsize_col = 10,
  fontsize_row = 10,
  cellheight = 25
)
```
## combined relapse avg exprs

```{r}
# Step 1: Filter the bulk matrix to keep only relapse samples
relapse_samples <- meta_data %>%
  filter(condition == "Relapse") %>%
  pull(sample)

relapse_matrix <- filtered_bulk_matrix[, relapse_samples, drop = FALSE]

# Step 2: Create an empty list to store the columns for each subtype with averages
relapse_extended_matrix <- as.data.frame(relapse_matrix)  # Start with the filtered matrix

# 1. Filter for ETV subtype columns
relapse_etv_matrix <- relapse_extended_matrix %>% dplyr::select(starts_with("ETV"))

# 2. Filter for LH subtype columns
relapse_lh_matrix <- relapse_extended_matrix %>% dplyr::select(starts_with("LH"))

# 3. Filter for NH subtype columns
relapse_nh_matrix <- relapse_extended_matrix %>% dplyr::select(starts_with("NH"))

# 4. Filter for PH subtype columns
relapse_ph_matrix <- relapse_extended_matrix %>% dplyr::select(starts_with("PH"))

# List of matrices and their respective average column names
relapse_matrices <- list(ETV = relapse_etv_matrix, LH = relapse_lh_matrix, NH = relapse_nh_matrix, PH = relapse_ph_matrix)
relapse_average_col_names <- c("ETV_avg", "LH_avg", "NH_avg", "PH_avg")

# Function to add average column to each matrix
add_avg_column <- function(matrix, avg_col_name, prefix) {
  matrix %>%
    dplyr::mutate(!!avg_col_name := rowMeans(dplyr::select(., starts_with(prefix)), na.rm = TRUE))
}

# Apply the function and assign each output to a separate variable
relapse_etv_matrix <- add_avg_column(relapse_etv_matrix, "ETV_avg", "ETV")
relapse_lh_matrix <- add_avg_column(relapse_lh_matrix, "LH_avg", "LH")
relapse_nh_matrix <- add_avg_column(relapse_nh_matrix, "NH_avg", "NH")
relapse_ph_matrix <- add_avg_column(relapse_ph_matrix, "PH_avg", "PH")

# Extract only the average columns from each matrix and rename them to subtype names
relapse_etv_avg <- relapse_etv_matrix %>% dplyr::select(ETV_avg)
relapse_lh_avg <- relapse_lh_matrix %>% dplyr::select(LH_avg)
relapse_nh_avg <- relapse_nh_matrix %>% dplyr::select(NH_avg)
relapse_ph_avg <- relapse_ph_matrix %>% dplyr::select(PH_avg)

# Combine the average columns into a single data frame, keeping row names intact
relapse_combined_avg_expression <- data.frame(
  ETV = relapse_etv_avg$ETV_avg,
  LH = relapse_lh_avg$LH_avg,
  NH = relapse_nh_avg$NH_avg,
  PH = relapse_ph_avg$PH_avg,
  row.names = rownames(relapse_etv_matrix)
)

# Check the structure to confirm the combined format
str(relapse_combined_avg_expression)

# Z-score normalization of each gene across the subtypes
relapse_zscore_matrix <- t(apply(relapse_combined_avg_expression, 1, function(x) (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)))

# Calculate the variance of each gene across subtypes
relapse_gene_variances <- apply(relapse_zscore_matrix, 1, var, na.rm = TRUE)

# Get the names of the top 500 most variable genes
relapse_top_500_genes <- names(sort(relapse_gene_variances, decreasing = TRUE))[1:500]

# Subset the z-score matrix to include only the top 500 genes
relapse_top_500_zscore_matrix <- relapse_zscore_matrix[relapse_top_500_genes, ]

# Filter the matrix to include only the row for gene "KIF9"
relapse_kif9_data <- relapse_zscore_matrix["GEM", , drop = FALSE]

# Plot the heatmap for the gene "KIF9"
pheatmap::pheatmap(
  relapse_kif9_data,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  main = "Heatmap for gene KIF9 on Relapse single cell",
  show_rownames = TRUE,
  show_colnames = TRUE,
  color = colorRampPalette(c("blue", "white", "red"))(50),
  fontsize_col = 10,
  fontsize_row = 10,
  cellheight = 25
)
```


# Task 4 Average Expression of StJudes Data

## Trying StJudes data

### Reading sample names

```{r}
patient_id <- readxl::read_excel("sample.subtype.pmid30643249.xlsx") %>% 
  rename("subtype" = "primary subtype")
```

```{r}
# Create a new column with transformed subtypes
patient_id <- patient_id %>%
  mutate(
    primary_subtype = case_when(
      subtype == "Ph-like" ~ "Ph",
      subtype == "ETV6-RUNX1-like" ~ "ETV6-RUNX1",
      subtype == "KMT2A-like" ~ "KMT2A",
      subtype == "ZNF384-like" ~ "ZNF384",
      TRUE ~ subtype  # Keep other subtypes as they are
    )
  )

# Display unique values to confirm changes
unique(patient_id$primary_subtype)
```


```{r}
stjudes <- read.delim("merged.Log2-FPKM_UQ.txt")
```

```{r}
# Load necessary library


# Remove "_D" and "_D1" suffixes from column names in `stjudes`
# Update `stjudes` column names to remove anything after the first underscore
names(stjudes) <- sub("_.*", "", names(stjudes))


stjudes_clean <- stjudes %>%
  dplyr::select(-GeneID)

# Step 2: Remove duplicate rows based on "Gene" column
stjudes_clean <- stjudes_clean %>%
  distinct(Gene, .keep_all = TRUE)

# Step 3: Set "Gene" column as row names and remove it from data frame
rownames(stjudes_clean) <- stjudes_clean$Gene
stjudes_clean <- stjudes_clean %>%
  dplyr::select(-Gene)


```


## Renaming

```{r}
rename_map <- patient_id %>%
  dplyr::select(patient, primary_subtype) %>%
  filter(patient %in% names(stjudes_clean))

# Create a named vector for renaming columns
rename_vector <- setNames(rename_map$primary_subtype, rename_map$patient)

# Convert `stjudes` to a matrix to allow duplicate column names
stjudes_matrix <- as.matrix(stjudes_clean)

# Apply the renaming directly, allowing duplicate names in the matrix
colnames(stjudes_matrix)[colnames(stjudes_matrix) %in% names(rename_vector)] <- rename_vector[colnames(stjudes_matrix)]

# Check if it matches your expectations
as.data.frame(head(stjudes_matrix))
```


## Doing the heatmap

```{r}
stjudes_df <- as.data.frame(stjudes_matrix)

# Step 2: Repair column names by appending unique suffixes to duplicates
colnames(stjudes_df) <- make.unique(colnames(stjudes_df), sep = "_")

# Step 3: Add row names as a column using `rownames_to_column`
stjudes_df <- stjudes_df %>% rownames_to_column("Gene")

```



```{r}
# Define subtypes to process
subtypes <- c("Ph", "Other", "ETV6-RUNX1", "PAX5alt", "Low hyperdiploid", "CRLF2(non-Ph-like)",
              "HLF", "PAX5 P80R", "MEF2D", "TCF3-PBX1", "High hyperdiploid", "DUX4", 
              "ZNF384", "KMT2A", "Near haploid", "iAMP21", "BCL2/MYC", "Low hypodiploid", 
              "IKZF1 N159Y", "NUTM1")

# Calculate average expression for each subtype
average_expression_list <- list()

for (subtype in subtypes) {
  subtype_avg <- stjudes_df %>%
    dplyr::select(Gene, contains(subtype)) %>%
    dplyr::mutate(avg_expression = rowMeans(dplyr::select(., -Gene), na.rm = TRUE)) %>%
    dplyr::select(Gene, avg_expression) %>%
    rename(!!paste0(subtype, "_avg") := avg_expression)
  
  average_expression_list[[subtype]] <- subtype_avg
}

# Step 4: Combine the averages for each subtype into a single data frame
stjudes_avg_expression <- Reduce(function(x, y) inner_join(x, y, by = "Gene"), average_expression_list)

# Display the combined data frame to confirm the structure
stjudes_avg_expression
```

## Plotting stjudes heatmap

```{r}
# Step 5: Z-score normalization across the averages
sjzscore_matrix <- t(apply(stjudes_avg_expression[, -1], 1, function(x) (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)))

# Step 6: Convert to a data frame and add gene names as row names
sjzscore_df <- as.data.frame(sjzscore_matrix)
rownames(sjzscore_df) <- stjudes_avg_expression$Gene

# Step 7: Plot the heatmap using the top genes by variance (e.g., top 50 or 100)
# Calculate variance and select top variable genes
sjgene_variances <- apply(sjzscore_df, 1, var, na.rm = TRUE)
sjtop_genes <- names(sort(sjgene_variances, decreasing = TRUE))[1:500]  # Adjust the number as needed

# Subset the z-score matrix to include only the top variable genes
sjtop_zscore_matrix <- sjzscore_df[sjtop_genes, ]

# Step 8: Plot the heatmap
pheatmap::pheatmap(
  sjtop_zscore_matrix,
  cluster_rows = T,  # Cluster genes based on Z-score patterns
  cluster_cols = F, # Keep the order of subtypes
  main = "Top 500 Variable Genes (Z-Score Normalized) of StJudes data Across Subtypes",
  show_rownames = TRUE,
  show_colnames = TRUE,
  color = colorRampPalette(c("blue", "white", "red"))(50), # Adjust color palette if needed
  fontsize_col = 9,
  fontsize_row = 1
)


```
## LH and NH

```{r}
# Convert Z-score matrix to a data frame if not already done
sjzscore_df <- as.data.frame(sjzscore_matrix)
rownames(sjzscore_df) <- stjudes_avg_expression$Gene

# Rank by `Low Hypodiploid`
ranked_low_hypodiploid <- sjzscore_df %>%
  arrange(desc(`Low hypodiploid_avg`))

# Rank by `Near Haploid`
ranked_near_haploid <- sjzscore_df %>%
  arrange(desc(`Near haploid_avg`))

```


### LH 

```{r}
# Select top 50 genes by variance from the ranked `Low Hypodiploid`
top_genes_low_hypodiploid <- head(rownames(ranked_low_hypodiploid), 50)
low_hypodiploid_zscore_matrix <- sjzscore_df[top_genes_low_hypodiploid, ]

# Plot the heatmap
pheatmap::pheatmap(
  low_hypodiploid_zscore_matrix,
  cluster_rows = TRUE,
  cluster_cols = FALSE,
  main = "Top 50 Variable Genes Ranked by Low Hypodiploid Z-Score",
  show_rownames = TRUE,
  show_colnames = TRUE,
  color = colorRampPalette(c("blue", "white", "red"))(50),
  fontsize_col = 7,
  fontsize_row = 9
)

```


## NH

```{r}
# Select top 50 genes by variance from the ranked `Near Haploid`
top_genes_near_haploid <- head(rownames(ranked_near_haploid), 50)
near_haploid_zscore_matrix <- sjzscore_df[top_genes_near_haploid, ]

# Plot the heatmap
pheatmap::pheatmap(
  near_haploid_zscore_matrix,
  cluster_rows = TRUE,
  cluster_cols = FALSE,
  main = "Top 50 Variable Genes Ranked by Near Haploid Z-Score",
  show_rownames = TRUE,
  show_colnames = TRUE,
  color = colorRampPalette(c("blue", "white", "red"))(50),
  fontsize_col = 7,
  fontsize_row = 9
)

```


```{r}
# Calculate variance and select top 1000 variable genes in St. Jude data
sjgene_variances <- apply(sjzscore_df, 1, var, na.rm = TRUE)
sjtop_genes_1000 <- names(sort(sjgene_variances, decreasing = TRUE))[1:1000]

# Subset the z-score matrix to include only the top 1000 genes
sjtop_zscore_matrix_1000 <- sjzscore_df[sjtop_genes_1000, ]
```


```{r}
# Rank the subset by `Low Hypodiploid`
ranked_low_hypodiploid_1000 <- sjtop_zscore_matrix_1000 %>%
  arrange(desc(`Low hypodiploid_avg`))

# Plot heatmap for top 1000 genes ranked by Low Hypodiploid Z-score
pheatmap::pheatmap(
  ranked_low_hypodiploid_1000,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  main = "Top 1000 Variable Genes Ranked by Low Hypodiploid Z-Score",
  show_rownames = FALSE,  # You may opt to hide row names for readability
  show_colnames = TRUE,
  color = colorRampPalette(c("blue", "white", "red"))(50),
  fontsize_col = 6,
  fontsize_row = 5
)

```

```{r}
# Rank the subset by `Near Haploid`
ranked_near_haploid_1000 <- sjtop_zscore_matrix_1000 %>%
  arrange(desc(`Near haploid_avg`))

# Plot heatmap for top 1000 genes ranked by Near Haploid Z-score
pheatmap::pheatmap(
  ranked_near_haploid_1000,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  main = "Top 1000 Variable Genes Ranked by Near Haploid Z-Score",
  show_rownames = FALSE,
  show_colnames = TRUE,
  color = colorRampPalette(c("blue", "white", "red"))(50),
  fontsize_col = 6,
  fontsize_row = 5
)

```

```{r}
# Export ranked data for Low Hypodiploid
write.csv(ranked_low_hypodiploid_1000, "top_1000_genes_ranked_by_low_hypodiploid.csv", row.names = TRUE)

# Export ranked data for Near Haploid
write.csv(ranked_near_haploid_1000, "top_1000_genes_ranked_by_near_haploid.csv", row.names = TRUE)

```


### Producing figure 1 part 2

```{r}
# Create a new DataFrame with column names cleaned of '_avg'
sjzscore_clean_df <- sjzscore_df
colnames(sjzscore_clean_df) <- gsub("_avg$", "", colnames(sjzscore_clean_df))

# Proceed with the original calculations and plots using the new DataFrame
# Calculate variance and select top 1000 variable genes in the cleaned data
sjgene_variances <- apply(sjzscore_clean_df, 1, var, na.rm = TRUE)
sjtop_genes_1000 <- names(sort(sjgene_variances, decreasing = TRUE))[1:1000]

# Subset the cleaned z-score matrix to include only the top 1000 genes
sjtop_zscore_matrix_1000 <- sjzscore_clean_df[sjtop_genes_1000, ]

# Rank the subset by `Low Hypodiploid`
ranked_low_hypodiploid_1000 <- sjtop_zscore_matrix_1000 %>%
  arrange(desc(`Low hypodiploid`))

# Rank the subset by `Near Haploid`
ranked_near_haploid_1000 <- sjtop_zscore_matrix_1000 %>%
  arrange(desc(`Near haploid`))

# Plot heatmap for top 1000 genes ranked by Low Hypodiploid Z-score
pheatmap::pheatmap(
  ranked_low_hypodiploid_1000,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  main = "Top 1000 Variable Genes Ranked by Low Hypodiploid Z-Score",
  show_rownames = FALSE,  # You may opt to hide row names for readability
  show_colnames = TRUE,
  color = colorRampPalette(c("blue", "white", "red"))(50),
  fontsize_col = 6,
  fontsize_row = 5
)



# Plot heatmap for top 1000 genes ranked by Near Haploid Z-score
pheatmap::pheatmap(
  ranked_near_haploid_1000,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  main = "Top 1000 Variable Genes Ranked by Near Haploid Z-Score",
  show_rownames = FALSE,
  show_colnames = TRUE,
  color = colorRampPalette(c("blue", "white", "red"))(50),
  fontsize_col = 6,
  fontsize_row = 5
)

# Create the directory if it doesn't exist
output_dir <- "/Users/cojulian/Desktop/Tzu_projects/Matt/Integrated_D_R_curtis/tasks_results/figure_1"
if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}

# Function to save pheatmap as SVG
save_pheatmap_svg <- function(pheatmap_obj, filename) {
  svg(filename)
  grid::grid.draw(pheatmap_obj$gtable)
  dev.off()
}

# Plot heatmap for top 1000 genes ranked by Low Hypodiploid Z-score and save as SVG
p1 <- pheatmap::pheatmap(
  ranked_low_hypodiploid_1000,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  main = "Top 1000 Variable Genes StJude's Ranked by Low Hypodiploid Z-Score",
  show_rownames = FALSE,  # You may opt to hide row names for readability
  show_colnames = TRUE,
  color = colorRampPalette(c("blue", "white", "red"))(50),
  fontsize_col = 6,
  fontsize_row = 5
)
save_pheatmap_svg(p1, file.path(output_dir, "stjudes_top1000_LowHypodiploidRanked.svg"))

# Plot heatmap for top 1000 genes ranked by Near Haploid Z-score and save as SVG
p2 <- pheatmap::pheatmap(
  ranked_near_haploid_1000,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  main = "Top 1000 Variable Genes in StJude's Ranked by Near Haploid Z-Score",
  show_rownames = FALSE,
  show_colnames = TRUE,
  color = colorRampPalette(c("blue", "white", "red"))(50),
  fontsize_col = 6,
  fontsize_row = 5
)
save_pheatmap_svg(p2, file.path(output_dir, "stjudes_top1000_NearHaploidRanked.svg"))

```

##### Task 4.2 scatterplot of GEM expression per patient

```{r}
library(dplyr)
library(ggplot2)
library(tidyr)

# Define the subtypes to keep
subtypes <- c("Ph", "Other", "ETV6-RUNX1", "PAX5alt", "Low hyperdiploid", "CRLF2(non-Ph-like)",
              "HLF", "PAX5 P80R", "MEF2D", "TCF3-PBX1", "High hyperdiploid", "DUX4", 
              "ZNF384", "KMT2A", "Near haploid", "iAMP21", "BCL2/MYC", "Low hypodiploid", 
              "IKZF1 N159Y", "NUTM1")

# Subset the matrix to keep only the specified subtypes
stjudes_subset <- stjudes_df %>%
  dplyr::select(Gene, contains(subtypes))

# Extract GEM gene expression
gem_expression <- stjudes_subset %>%
  filter(Gene == "GEM") %>%
  pivot_longer(cols = -Gene, names_to = "Patient", values_to = "Expression")

# Extract patient subtypes
patient_subtypes <- patient_id %>%
  dplyr::select(patient, primary_subtype) %>%
  filter(patient %in% gem_expression$Patient)

# Merge GEM expression with patient subtypes
gem_expression <- gem_expression %>%
  left_join(patient_subtypes, by = c("Patient" = "patient"))

# Plot GEM gene expression with individual patients on the x-axis
ggplot(gem_expression, aes(x = Patient, y = Expression, color = primary_subtype)) +
  geom_jitter(width = 0.2) +
  labs(title = "GEM Gene Expression by Patient",
       x = "Patient",
       y = "GEM Gene Expression (Log2 FPKM)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

```{r}
library(dplyr)
library(tidyr)

library(dplyr)
library(ggplot2)
library(tidyr)

# Step 1: Extract GEM gene expression
gem_expression <- stjudes_df %>%
  filter(Gene == "GEM") %>%  # Keep only the row for the GEM gene
  pivot_longer(cols = -Gene, names_to = "patient", values_to = "GEM_expression")  # Reshape to long format

# Step 2: Merge with patient_id to add primary_subtype
gem_expression_with_subtype <- gem_expression %>%
  left_join(patient_id, by = "patient")  # Merge with patient_id

# Step 3: Extract subtype from patient IDs
gem_expression_with_subtype <- gem_expression_with_subtype %>%
  mutate(subtype = gsub("(_\\d+)$", "", patient))  # Remove suffixes like _1, _2, etc.

# Step 4: Plot boxplots with scatterplot overlay
ggplot(gem_expression_with_subtype, aes(x = subtype, y = GEM_expression, fill = subtype)) +
  geom_boxplot(outlier.shape = NA) +  # Boxplot without outliers (we'll add them as dots)
  geom_jitter(width = 0.2, alpha = 0.5, color = "black") +  # Add jittered points
  labs(
    title = "GEM Gene Expression by Subtype",
    x = "Subtype",
    y = "GEM Gene Expression (Log2 FPKM)"
  ) +
  theme_minimal() +
 theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 0.5),
        legend.position="none")

```

```{r}
library(dplyr)
library(ggplot2)
library(tidyr)

# Step 1: Extract GEM gene expression
gem_expression <- stjudes_df %>%
  filter(Gene == "GEM") %>%  # Keep only the row for the GEM gene
  pivot_longer(cols = -Gene, names_to = "patient", values_to = "GEM_expression")  # Reshape to long format

# Step 2: Merge with patient_id to add primary_subtype
gem_expression_with_subtype <- gem_expression %>%
  left_join(patient_id, by = "patient")  # Merge with patient_id

# Step 3: Extract subtype from patient IDs
gem_expression_with_subtype <- gem_expression_with_subtype %>%
  mutate(subtype = gsub("(_\\d+)$", "", patient))  # Remove suffixes like _1, _2, etc.

# Step 4: Calculate median GEM expression for each subtype
subtype_medians <- gem_expression_with_subtype %>%
  group_by(subtype) %>%
  summarise(median_expression = median(GEM_expression, na.rm = TRUE)) %>%
  arrange(desc(median_expression))  # Sort subtypes by median expression (highest to lowest)

# Step 5: Reorder the subtype factor based on median expression
gem_expression_with_subtype <- gem_expression_with_subtype %>%
  mutate(subtype = factor(subtype, levels = subtype_medians$subtype))

# Step 6: Plot boxplots with scatterplot overlay, ordered by median expression
ggplot(gem_expression_with_subtype, aes(x = subtype, y = GEM_expression, fill = subtype)) +
  geom_boxplot(outlier.shape = NA) +  # Boxplot without outliers (we'll add them as dots)
  geom_jitter(width = 0.2, alpha = 0.5, color = "black") +  # Add jittered points
  labs(
    title = "GEM Gene Expression by Subtype (Ordered by Median Expression)",
    x = "Subtype",
    y = "GEM Gene Expression (Log2 FPKM)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 0.5),
    legend.position = "none"  # Remove legend
  )
```

### Loop for different genes

```{r}
library(dplyr)
library(ggplot2)
library(tidyr)

# Define the list of genes
genes_of_interest <- c("GEM", "INHBE", "RASL11A", "CYP7B1", "NFKBIE", "ZBTB10", "CTSB", "OLAH", "FAM76B")

# Loop through each gene
for (gene in genes_of_interest) {
  
  # Step 1: Extract gene expression
  gene_expression <- stjudes_df %>%
    filter(Gene == gene) %>%  # Keep only the row for the current gene
    pivot_longer(cols = -Gene, names_to = "patient", values_to = "expression")  # Reshape to long format
  
  # Step 2: Merge with patient_id to add primary_subtype
  gene_expression_with_subtype <- gene_expression %>%
    left_join(patient_id, by = "patient")  # Merge with patient_id
  
  # Step 3: Extract subtype from patient IDs
  gene_expression_with_subtype <- gene_expression_with_subtype %>%
    mutate(subtype = gsub("(_\\d+)$", "", patient))  # Remove suffixes like _1, _2, etc.
  
  # Step 4: Calculate median expression for each subtype
  subtype_medians <- gene_expression_with_subtype %>%
    group_by(subtype) %>%
    summarise(median_expression = median(expression, na.rm = TRUE)) %>%
    arrange(median_expression)  # Sort subtypes by median expression in ASCENDING order
  
  # Step 5: Reorder the subtype factor based on median expression (ascending)
  gene_expression_with_subtype <- gene_expression_with_subtype %>%
    mutate(subtype = factor(subtype, levels = subtype_medians$subtype))
  
  # Step 6: Plot boxplots with scatterplot overlay, ordered by median expression
  p <- ggplot(gene_expression_with_subtype, aes(x = subtype, y = expression, fill = subtype)) +
    geom_boxplot(outlier.shape = NA) +  # Boxplot without outliers (we'll add them as dots)
    geom_jitter(width = 0.2, alpha = 0.5, color = "black") +  # Add jittered points
    labs(
      title = paste0(gene, " Gene Expression by Subtype (Ordered by Median Expression)"),
      x = "Subtype",
      y = paste0(gene, " Gene Expression (Log2 FPKM)")
    ) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),  # Adjust text angle and alignment
      plot.margin = unit(c(1, 1, 1, 1.5), "cm"),  # Add margin to avoid text cutoff
      legend.position = "none"  # Remove legend
    )
  
  # Display the plot
  print(p)
  
  ggsave(
  filename = file.path("/Users/cojulian/Desktop/Tzu_projects/Matt/Integrated_D_R_curtis/tasks_results/median_plots", 
                       paste0(gene, "_median_expression_stjude.png")),  # Full path to save location
  plot = p,  # The plot to save
  width = 12,  # Width of the plot (in inches)
  height = 8,  # Height of the plot (in inches)
  dpi = 600,  # High resolution (600 DPI)
  bg = "white"  # Background color (optional, ensures white background)
)
}
```


LH
NH

in the slides
mention what matt wanted to see both bulk and single cell, row and columnwise


### Ascending meand or z-score

```{r}
gem_expression_with_subtype <- gem_expression_with_subtype %>%
  group_by(subtype) %>% 
  mutate(GEM_zscore = (GEM_expression - mean(GEM_expression, na.rm = TRUE)) /
                        sd(GEM_expression, na.rm = TRUE)) %>%
  ungroup()


ggplot(gem_expression_with_subtype, aes(x = subtype, y = GEM_zscore, fill = subtype)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.5, color = "black") +
  labs(
    title = "GEM Gene Expression Z-Scores by Subtype",
    x = "Subtype",
    y = "GEM Gene Expression (Z-Score)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 0.5),
    legend.position = "none"
  )

```

```{r}
# Calculate median z-score for each subtype
subtype_medians <- gem_expression_with_subtype %>%
  group_by(subtype) %>%
  summarise(median_z = median(GEM_zscore, na.rm = TRUE)) %>%
  arrange(median_z)  # ascending order

# Reorder the factor 'subtype' based on the ascending median z-score
gem_expression_with_subtype <- gem_expression_with_subtype %>%
  mutate(subtype = factor(subtype, levels = subtype_medians$subtype))

# Plot the reordered data
ggplot(gem_expression_with_subtype, aes(x = subtype, y = GEM_zscore, fill = subtype)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.5, color = "black") +
  labs(
    title = "GEM Gene Expression Z-Scores by Subtype (Ascending Order)",
    x = "Subtype",
    y = "GEM Gene Expression (Z-Score)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 0.5),
    legend.position = "none"
  )
```

## Task 4.3.
#### Trying it rowise

##### Deprecated
```{r}
# z_scores <- stjudes_df %>%
#   column_to_rownames("Gene") %>%
#   as.matrix() %>%
#   apply(1, function(x) if (sd(x, na.rm = TRUE) == 0) rep(NA, length(x)) else (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)) %>%
#   t() %>%
#   as.data.frame() %>%
#   rownames_to_column("Gene")
```

```{r}

# Update column names to replace "CRLF2(non-Ph-like)" with "CRLF2_nonPhlike"
colnames(stjudes_df) <- gsub("CRLF2\\(non-Ph-like\\)", "CRLF2_nonPhlike", colnames(stjudes_df))

# Verify the updated column names
print(grep("CRLF2_nonPhlike", colnames(stjudes_df), value = TRUE))

# Define subtypes to process
subtypes <- c("Ph", "Other", "ETV6-RUNX1", "PAX5alt", "Low hyperdiploid", "CRLF2_nonPhlike",
              "HLF", "PAX5 P80R", "MEF2D", "TCF3-PBX1", "High hyperdiploid", "DUX4", 
              "ZNF384", "KMT2A", "Near haploid", "iAMP21", "BCL2/MYC", "Low hypodiploid", 
              "IKZF1 N159Y", "NUTM1")

# Calculate average expression for each subtype
average_expression_list <- list()

for (subtype in subtypes) {
  # Select columns that match the subtype (including those with _1, _2, etc.)
  subtype_cols <- grep(paste0("^", subtype, "(_\\d+)?$"), colnames(stjudes_df), value = TRUE)
  
  # Calculate the average expression for the subtype
  subtype_avg <- stjudes_df %>%
    dplyr::select(Gene, all_of(subtype_cols)) %>%
    dplyr::mutate(avg_expression = rowMeans(dplyr::select(., -Gene), na.rm = TRUE)) %>%
    dplyr::select(Gene, avg_expression) %>%
    rename(!!paste0(subtype, "_avg") := avg_expression)
  
  # Add the result to the list
  average_expression_list[[subtype]] <- subtype_avg
}

# Combine the averages for each subtype into a single data frame
stjudes_avg_expression <- Reduce(function(x, y) inner_join(x, y, by = "Gene"), average_expression_list)

# Check the result
head(stjudes_avg_expression)
names(stjudes_avg_expression)
```
### Step 2: Generate Z-Scores for Each Gene Across Subtypes


```{r}
# Step 2: Generate Z-scores for each gene across subtypes
zscore_matrix <- stjudes_avg_expression %>%
  column_to_rownames("Gene") %>%  # Set 'Gene' as row names
  as.matrix() %>%  # Convert to a matrix
  apply(1, function(x) {
    if (sd(x, na.rm = TRUE) == 0) {
      rep(NA, length(x))  # Handle genes with zero standard deviation
    } else {
      (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)  # Compute Z-score
    }
  }) %>%
  t()  # Transpose back to original shape

# Convert back to a dataframe and restore the 'Gene' column
z_scores <- zscore_matrix %>%
  as.data.frame() %>%
  rownames_to_column("Gene")

# Check the result
head(z_scores)
```

### Step 3: Identify Genes with Z-Score > 2 for Both LH and NH

```{r}
# Step 3: Determine which genes have a Z-score > 2 for LH AND NH
# Define LH and NH subtypes
lh_subtype <- "Low hypodiploid_avg"  # Replace with the actual subtype name if different
nh_subtype <- "Near haploid_avg"      # Replace with the actual subtype name if different

# Filter genes with Z-score > 2 for both LH and NH
genes_with_high_zscore <- z_scores %>%
  filter(
    !!sym(lh_subtype) > 2 &  # Z-score > 2 for LH
    !!sym(nh_subtype) > 2    # Z-score > 2 for NH
  )

# Check the result
genes_with_high_zscore
```


#### Heatmap of zscore filtered > 2

```{r}
# Set 'Gene' as row names and convert to a numeric matrix
heatmap_data <- genes_with_high_zscore %>%
  column_to_rownames("Gene") %>%
  as.matrix()

# Generate the heatmap
pheatmap::pheatmap(
  heatmap_data,
  cluster_rows = T,       # Do not cluster rows (genes)
  cluster_cols = T,       # Do not cluster columns (subtypes)
  main = "Genes with High Z-Scores in StJude's Dataset",  # Heatmap title
  show_rownames = TRUE,       # Show gene names (set to FALSE if too many genes)
  show_colnames = TRUE,       # Show subtype names
  color = colorRampPalette(c("blue", "white", "red"))(50),  # Color gradient
  fontsize_col = 8,           # Font size for column names
  fontsize_row = 6            # Font size for row names
)
```
#### Now doing AND/OR and filtering

### Step 3: Identify Genes with Z-Score > 2 for Both LH and NH

```{r}
# Step 3: Determine which genes have a Z-score > 2 for LH AND NH
# Define LH and NH subtypes
lh_subtype <- "Low hypodiploid_avg"  # Replace with the actual subtype name if different
nh_subtype <- "Near haploid_avg"      # Replace with the actual subtype name if different

# Filter genes with Z-score > 2 for both LH and NH
genes_with_high_zscore <- z_scores %>%
  filter(
    !!sym(lh_subtype) > 2 &  # Z-score > 2 for LH
    !!sym(nh_subtype) > 2    # Z-score > 2 for NH
  )

# Check the result
genes_with_high_zscore
```



# REMOVING LOW EXPRESSING

```{r}
# Define subtypes
subtypes <- c("Ph", "Other", "ETV6RUNX1", "PAX5alt", "Lowhyperdiploid", "CRLF2_nonPhlike",
              "HLF", "PAX5P80R", "MEF2D", "TCF3PBX1", "Highhyperdiploid", "DUX4", 
              "ZNF384", "KMT2A", "Nearhaploid", "iAMP21", "BCL2MYC", "Lowhypodiploid", 
              "IKZF1N159Y", "NUTM1")

# Function to check if a gene is expressed in a majority of a subtype's samples
check_majority_positive <- function(expression_values) {
  sum(expression_values > 0, na.rm = TRUE) > (length(expression_values) / 2)
}

# Step 1: Create a logical matrix indicating whether each gene is expressed per subtype
subtype_expression_matrix <- sapply(subtypes, function(subtype) {
  # Find matching columns for the subtype
  subtype_cols <- grep(paste0("^", subtype, "(_\\d+)?$"), colnames(stjudes_df), value = TRUE)
  
  # Skip if no matching columns were found
  if (length(subtype_cols) == 0) return(rep(FALSE, nrow(stjudes_df)))
  
  # Apply function to check majority expression for each gene
  apply(stjudes_df[subtype_cols], 1, check_majority_positive)
})

# Step 2: Filter genes where at least one subtype has a majority positive expression
filtered_stjudes_df <- stjudes_df[rowSums(subtype_expression_matrix) > 0, ]

# Check results
dim(filtered_stjudes_df)  # Should be fewer rows than the original dataset
```

#### Step 1: Compute Average Expression per Subtype in filtered_stjudes_df

```{r}
library(dplyr)

# Update column names to replace "CRLF2(non-Ph-like)" with "CRLF2_nonPhlike"
colnames(filtered_stjudes_df) <- gsub("CRLF2\\(non-Ph-like\\)", "CRLF2_nonPhlike", colnames(filtered_stjudes_df))

# Verify the updated column names
print(grep("CRLF2_nonPhlike", colnames(filtered_stjudes_df), value = TRUE))

# Define subtypes to process
subtypes <- c("Ph", "Other", "ETV6-RUNX1", "PAX5alt", "Low hyperdiploid", "CRLF2_nonPhlike",
              "HLF", "PAX5 P80R", "MEF2D", "TCF3-PBX1", "High hyperdiploid", "DUX4", 
              "ZNF384", "KMT2A", "Near haploid", "iAMP21", "BCL2/MYC", "Low hypodiploid", 
              "IKZF1 N159Y", "NUTM1")

# Calculate average expression for each subtype
lowexp_filt_avg_expression_list <- list()

for (subtype in subtypes) {
  # Select columns that match the subtype (including those with _1, _2, etc.)
  subtype_cols <- grep(paste0("^", subtype, "(_\\d+)?$"), colnames(filtered_stjudes_df), value = TRUE)
  
  # Calculate the average expression for the subtype
  subtype_avg <- filtered_stjudes_df %>%
    dplyr::select(Gene, all_of(subtype_cols)) %>%
    dplyr::mutate(avg_expression = rowMeans(dplyr::select(., -Gene), na.rm = TRUE)) %>%
    dplyr::select(Gene, avg_expression) %>%
    rename(!!paste0(subtype, "_avg") := avg_expression)
  
  # Add the result to the list
  lowexp_filt_avg_expression_list[[subtype]] <- subtype_avg
}

# Combine the averages for each subtype into a single data frame
lowexp_filt_avg_expression <- Reduce(function(x, y) inner_join(x, y, by = "Gene"), lowexp_filt_avg_expression_list)

# Check the result
head(lowexp_filt_avg_expression)
names(lowexp_filt_avg_expression)
```

####  Step 2: Generate Z-Scores for Each Gene Across Subtypes

```{r}
library(tibble)  # Ensure tibble functions are available

# Generate Z-scores for each gene across subtypes
lowexp_filt_zscore_matrix <- lowexp_filt_avg_expression %>%
  column_to_rownames("Gene") %>%  # Set 'Gene' as row names
  as.matrix() %>%  # Convert to a matrix
  apply(1, function(x) {
    if (sd(x, na.rm = TRUE) == 0) {
      rep(NA, length(x))  # Handle genes with zero standard deviation
    } else {
      (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)  # Compute Z-score
    }
  }) %>%
  t()  # Transpose back to original shape

# Convert back to a dataframe and restore the 'Gene' column
lowexp_filt_z_scores <- lowexp_filt_zscore_matrix %>%
  as.data.frame() %>%
  rownames_to_column("Gene")

# Check the result
head(lowexp_filt_z_scores)

```
####  Step 3: Identify Genes with Z-Score > 2 for Both LH and NH

```{r}
# Define LH and NH subtypes
lh_subtype <- "Low hypodiploid_avg"  # Replace with actual subtype name if different
nh_subtype <- "Near haploid_avg"      # Replace with actual subtype name if different

# Filter genes with Z-score > 2 in either LH or NH (OR) as well as both (AND)
lowexp_filt_genes_with_high_zscore <- lowexp_filt_z_scores %>%
  filter(
    (!!sym(lh_subtype) >= 2.5) |  # OR: Z-score > 2.5 for LH
    (!!sym(nh_subtype) >= 2.5)    # OR: Z-score > 2.5 for NH
  )

# Check the result
lowexp_filt_genes_with_high_zscore
```

##### Option 2: The 797 AND/OR genes in Z-score
```{r}
# Keep only genes that are in `lowexp_filt_genes_with_high_zscore`
lowexp_filt_highzscore_log2fpkm <- lowexp_filt_avg_expression %>%
  filter(Gene %in% lowexp_filt_genes_with_high_zscore$Gene)

# Check dimensions (should match number of genes in `lowexp_filt_genes_with_high_zscore`)
dim(lowexp_filt_highzscore_log2fpkm)

# Save filtered log2FPKM dataframe
#write.csv(lowexp_filt_highzscore_log2fpkm, "Filtered_HighZscore_Log2FPKM.csv", row.names = FALSE)

```

##### Option 2: The 797 AND/OR genes in Z-score

```{r}
# Keep genes where at least one subtype has log2FPKM >= 1
lowexp_filt_highzscore_log2fpkm_filt <- lowexp_filt_highzscore_log2fpkm %>%
  rowwise() %>%
  filter(any(across(where(is.numeric), ~ .x >= 1))) %>%
  ungroup()

# Check dimensions (should be fewer than previous dataset)
dim(lowexp_filt_highzscore_log2fpkm_filt)

```

```{r}
library(pheatmap)

# Set 'Gene' as row names and convert to a numeric matrix
lowexp_filt_heatmap_data <- lowexp_filt_highzscore_log2fpkm_filt %>%
  column_to_rownames("Gene") %>%
  as.matrix()

# Generate the heatmap
pheatmap::pheatmap(
  lowexp_filt_heatmap_data,
  cluster_rows = T,       # Cluster rows (genes)
  cluster_cols = T,       # Cluster columns (subtypes)
  main = "Genes with High Z-Scores & log2FPKM  1 (Filtered for Low Expression) in log2FPKM",  # Heatmap title
  show_rownames = TRUE,       # Show gene names (set to FALSE if too many genes)
  show_colnames = TRUE,       # Show subtype names
  color = colorRampPalette(c("blue", "white", "red"))(50),  # Color gradient
  fontsize_col = 8,           # Font size for column names
  fontsize_row = 6,
  legend = TRUE,
  legend_labels = c("title")# Font size for row names
)

```

```{r}
library(pheatmap)

# Set 'Gene' as row names and convert to a numeric matrix
lowexp_filt_heatmap_zscore_data <- lowexp_filt_genes_with_high_zscore %>%
  column_to_rownames("Gene") %>%
  as.matrix()

# Generate the heatmap
pheatmap::pheatmap(
  lowexp_filt_heatmap_zscore_data,
  cluster_rows = T,       # Cluster rows (genes)
  cluster_cols = T,       # Cluster columns (subtypes)
  main = "Genes with High Z-Scores (Filtered for Low Expression) in Z-Scores",  # Heatmap title
  show_rownames = TRUE,       # Show gene names (set to FALSE if too many genes)
  show_colnames = TRUE,       # Show subtype names
  color = colorRampPalette(c("blue", "white", "red"))(50),  # Color gradient
  fontsize_col = 8,           # Font size for column names
  fontsize_row = 6
  )

```
### Doing GSEA of 167 genes

```{r}
# Load required packages
library(msigdbr)
library(fgsea)
library(dplyr)
library(stringr)

# Check available species in msigdbr (for Homo sapiens)
msigdbr_show_species()

# Retrieve Hallmark (H) and C6 Oncogenic pathways for Homo sapiens
msigdbr_human_hallmarks <- msigdbr(species = "Homo sapiens", category = "H")
msigdbr_human_oncogenic <- msigdbr(species = "Homo sapiens", category = "C6")

```

```{r}
# Prepare Hallmark pathway list for fgsea
fgsea_hallmark_sets <- msigdbr_human_hallmarks %>%
  mutate(gs_name = str_remove(gs_name, "HALLMARK_")) %>%  # Remove prefix for readability
  split(x = .$gene_symbol, f = .$gs_name)

# Prepare C6 Oncogenic Signature pathway list for fgsea
fgsea_oncogenic_sets <- msigdbr_human_oncogenic %>%
  split(x = .$gene_symbol, f = .$gs_name)

```


```{r}
# Compute mean log2FPKM expression for each gene across all subtypes
gene_ranks <- lowexp_filt_highzscore_log2fpkm_filt %>%
  rowwise() %>%
  mutate(mean_expression = mean(c_across(where(is.numeric)), na.rm = TRUE)) %>%
  ungroup() %>%
  dplyr::select(Gene, mean_expression) %>%
  arrange(desc(mean_expression)) %>%  # Sort in descending order
  deframe()  # Convert to named vector for fgsea

```


```{r}
# Run GSEA for Hallmark pathways
fgseaRes_hallmark <- fgsea(pathways = fgsea_hallmark_sets, stats = gene_ranks)

# Run GSEA for C6 Oncogenic Signatures
fgseaRes_oncogenic <- fgsea(pathways = fgsea_oncogenic_sets, stats = gene_ranks)

```

```{r}
# Tidy up Hallmark GSEA results
fgseaResTidy_hallmark <- fgseaRes_hallmark %>%
  mutate(abbrev_pathway = abbreviate(pathway)) %>%
  relocate(abbrev_pathway, .after = pathway) %>%
  as_tibble() %>%
  arrange(padj)  # Sort by adjusted p-value

# Tidy up C6 Oncogenic GSEA results
fgseaResTidy_oncogenic <- fgseaRes_oncogenic %>%
  mutate(abbrev_pathway = abbreviate(pathway)) %>%
  relocate(abbrev_pathway, .after = pathway) %>%
  as_tibble() %>%
  arrange(padj)  # Sort by adjusted p-value

# View top 10 enriched Hallmark pathways
fgseaResTidy_hallmark %>% head(10)

# View top 10 enriched Oncogenic pathways
fgseaResTidy_oncogenic %>% head(10)

```
```{r}
library(ggplot2)

fgseaResTidy_hallmark %>%
  mutate(NES2 = ifelse(NES > 0, "Positive", "Negative")) %>%
  arrange(padj) %>%
  head(20) %>%
  ggplot(aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill = NES2)) +
  scale_fill_manual(breaks = c("Positive", "Negative"), values = c("#61BB6F", "#EC7012")) +
  theme(axis.text.y = element_text(size = 6)) +
  coord_flip() +
  labs(x = "Pathway", y = "Normalized Enrichment Score",
       title = "Top 20 Hallmark Pathways (High Z-Score St Jude's Bulk-RNAseq Genes)")

```



# Task 5 Overlapping, LH only, NH only

## Filtering

```{r}
selected_genes <- readxl::read_excel("St Jude Overlap_LH_NH_110824.xlsx")
```

```{r}
# Step 1: Extract unique gene sets from the provided table `selected_genes`
nh_only_genes <- unique(selected_genes$`NH only (178)`) %>% na.omit()
lh_only_genes <- unique(selected_genes$`LH (88)`) %>% na.omit()
overlap_genes <- unique(selected_genes$`Overlap (52)`) %>% na.omit()

# Step 2: Combine genes in the specified order for the heatmap
# NH-only genes at the top, followed by overlap genes, and then LH-only genes
ordered_genes <- c(nh_only_genes, overlap_genes, lh_only_genes)

# Step 3: Filter `ordered_genes` to only include genes that are present in `zscore_matrix`
# This ensures we're only using genes that actually have data in the matrix
valid_genes <- ordered_genes[ordered_genes %in% rownames(zscore_matrix)]

# Step 4: Subset `zscore_matrix` to include only `valid_genes` and in the specified order
# `drop = FALSE` keeps it as a matrix if there's only one gene
filtered_zscore_matrix <- zscore_matrix[valid_genes, , drop = FALSE]  
# Create an annotation data frame to label each gene
gene_group <- rep(NA, length(valid_genes))
gene_group[valid_genes %in% nh_only_genes] <- "NH only"
gene_group[valid_genes %in% overlap_genes] <- "Overlap"
gene_group[valid_genes %in% lh_only_genes] <- "LH only"

annotation <- data.frame(Group = gene_group)
rownames(annotation) <- valid_genes

# Plot the heatmap with the annotation
pheatmap(
  filtered_zscore_matrix,
  annotation_row = annotation,        # Add row annotation for gene groups
  cluster_rows = F,               # Keep the gene order as specified
  cluster_cols = F,               # No clustering for subtypes
  main = "Z-Score Heatmap with NH, Overlap, and LH Groups",
  show_rownames = TRUE,
  show_colnames = TRUE,
  color = colorRampPalette(c("blue", "white", "red"))(50),
  fontsize_col = 8,
  fontsize_row = 6
)

```

## Anotteing 3

```{r}
# Separate genes by groups
nh_only_genes <- unique(selected_genes$`NH only (178)`) %>% na.omit()
overlap_genes <- unique(selected_genes$`Overlap (52)`) %>% na.omit()
lh_only_genes <- unique(selected_genes$`LH (88)`) %>% na.omit()

# Filter z-score matrix for each gene group
filtered_zscore_matrix_nh_only <- filtered_zscore_matrix[rownames(filtered_zscore_matrix) %in% nh_only_genes, ]
filtered_zscore_matrix_overlap <- filtered_zscore_matrix[rownames(filtered_zscore_matrix) %in% overlap_genes, ]
filtered_zscore_matrix_lh_only <- filtered_zscore_matrix[rownames(filtered_zscore_matrix) %in% lh_only_genes, ]


filtered_zscore_matrix_nh_only <- as.data.frame(filtered_zscore_matrix_nh_only) %>%
  arrange(desc(`Near haploid_avg`)) %>% 
  as.matrix()


filtered_zscore_matrix_lh_only <- as.data.frame(filtered_zscore_matrix_lh_only) %>%
  arrange(desc(`Low hyperdiploid_avg`)) %>% 
  as.matrix()



# Plot heatmap for NH only genes
pheatmap(
  filtered_zscore_matrix_nh_only,
  cluster_rows = F,               # Cluster genes within the NH only group
  cluster_cols = FALSE,               # No clustering for subtypes
  main = "Z-Score Heatmap for NH Only Genes",
  show_rownames = TRUE,
  show_colnames = TRUE,
  color = colorRampPalette(c("blue", "white", "red"))(50),
  fontsize_col = 8,
  fontsize_row = 6
)

# Plot heatmap for Overlap genes
pheatmap(
  filtered_zscore_matrix_overlap,
  cluster_rows = TRUE,               # Cluster genes within the Overlap group
  cluster_cols = FALSE,               # No clustering for subtypes
  main = "Z-Score Heatmap for Overlap Genes",
  show_rownames = TRUE,
  show_colnames = TRUE,
  color = colorRampPalette(c("blue", "white", "red"))(50),
  fontsize_col = 8,
  fontsize_row = 6
)

# Plot heatmap for LH only genes
pheatmap(
  filtered_zscore_matrix_lh_only,
  cluster_rows = F,               # Cluster genes within the LH only group
  cluster_cols = FALSE,               # No clustering for subtypes
  main = "Z-Score Heatmap for LH Only Genes",
  show_rownames = TRUE,
  show_colnames = TRUE,
  color = colorRampPalette(c("blue", "white", "red"))(50),
  fontsize_col = 8,
  fontsize_row = 6
)

```

# Task 6  Pharmacotypes (Quartile plots of LC50)

```{r}
ic50_drugs_gene <- read.csv("imputed_lc50_data.csv") %>% 
  dplyr::select(-c(".id", ".imp"))
rnaseq_drugs_list <- read.csv("pharmacotyping_ped_rnaseq_fpkm_ALLids_0823.csv")
```

### Colapsing the imputations

```{r}
# Collapse imputations into one average row per patient
# Collapse imputations and overwrite the original data
ic50_drugs_gene_list <- ic50_drugs_gene %>%
    group_by(subject_id) %>%
    summarise(across(starts_with("CHZ"):Vorinostat, mean, na.rm = TRUE)) %>%
    ungroup()

######
# Here remember that I put 'ic50_drugs_gene' for using all 10 imputations and 'ic50_drugs_gene_list' for using their avg
#####

# Inspect the resulting dataframe
print(head(ic50_drugs_gene_list))
```


## Task 6.1: Rank Patient Samples Based on Expression of Each Gene


```{r}
# Define the list of target genes
target_genes <- c("GEM", "INHBE", "RASL11A", "CYP7B1", "NFKBIE", 
                  "ZBTB10", "CTSB", "OLAH", "FAM76B")


# Extract and process GEM expression data
if ("GEM" %in% rnaseq_drugs_list$GeneName) {
    gene_expression <- rnaseq_drugs_list[rnaseq_drugs_list$GeneName == "GEM", ]
    expression_values <- gene_expression[, !(names(gene_expression) %in% c("GeneID", "GeneName")), drop = FALSE]
    expression_df <- as.data.frame(t(expression_values))
    colnames(expression_df) <- "Expression"
    expression_df$Patient <- rownames(expression_df)
} else {
    stop("Gene GEM not found in the dataset.")
}
```


```{r}
# Define quartiles for GEM expression
expression_df$Quartile <- cut(
    expression_df$Expression,
    breaks = quantile(expression_df$Expression, probs = seq(0, 1, 0.25), na.rm = TRUE),
    labels = c("Q1", "Q2", "Q3", "Q4"),
    include.lowest = TRUE
)
```



```{r}
# Merge expression data with IC50 data
merged_df <- merge(expression_df, ic50_drugs_gene, by.x = "Patient", by.y = "subject_id")
```


```{r}
# Step 4: Perform Kruskal-Wallis and Pairwise Wilcoxon Tests for Each Drug

# Prepare an empty data frame to store pairwise test results
pairwise_results <- data.frame()

# Pivot to long format for easier manipulation
drug_data_long <- merged_df %>%
    pivot_longer(cols = starts_with("CHZ"):Vorinostat, names_to = "Drug", values_to = "LC50")

# Loop over each drug to perform Kruskal-Wallis and pairwise Wilcoxon tests
for (drug in unique(drug_data_long$Drug)) {
    # Filter data for the specific drug
    drug_data <- drug_data_long %>% filter(Drug == drug)
    
    # Perform Kruskal-Wallis test
    kw_test <- kruskal_test(data = drug_data, LC50 ~ Quartile)
    
    # If the Kruskal-Wallis test is significant, proceed with pairwise Wilcoxon tests
    if (kw_test$p < 0.05) {
        # Conduct pairwise Wilcoxon test and adjust for multiple testing
        pw_test <- drug_data %>%
            pairwise_wilcox_test(LC50 ~ Quartile, p.adjust.method = "fdr") %>%
            mutate(Drug = drug)  # Add the drug name to the results
        
        # Append results to the pairwise_results data frame
        pairwise_results <- bind_rows(pairwise_results, pw_test)
    }
}

# Step 5: Finalize and Display Results
# Clean up the results, focusing on relevant columns
pairwise_results <- pairwise_results %>%
    dplyr::select(Drug, group1, group2, p, p.adj, p.adj.signif) %>%
    arrange(Drug, p.adj)

# Display the pairwise results for interpretation
print(pairwise_results)
```

## All genes across all drugs AVERAGE

```{r}
# Define the list of target genes
target_genes <- c("GEM", "INHBE", "RASL11A", "CYP7B1", "NFKBIE", 
                  "ZBTB10", "CTSB", "OLAH", "FAM76B")

# Initialize a list to store pairwise results for each gene
all_pairwise_results <- list()

# Loop through each gene to perform statistical tests
for (gene in target_genes) {
  # Extract and process expression data for the current gene
  if (gene %in% rnaseq_drugs_list$GeneName) {
    gene_expression <- rnaseq_drugs_list[rnaseq_drugs_list$GeneName == gene, ]
    expression_values <- gene_expression[, !(names(gene_expression) %in% c("GeneID", "GeneName")), drop = FALSE]
    expression_df <- as.data.frame(t(expression_values))
    colnames(expression_df) <- "Expression"
    expression_df$Patient <- rownames(expression_df)
  } else {
    stop(paste("Gene", gene, "not found in the dataset."))
  }

  # Define quartiles for the current gene expression
  quantiles <- quantile(expression_df$Expression, probs = seq(0, 1, 0.25), na.rm = TRUE)
  if (length(unique(quantiles)) < 5) {
    message(paste("Quantiles are not unique for gene", gene, "- skipping."))
    next
  }
  
  expression_df$Quartile <- cut(
    expression_df$Expression,
    breaks = quantiles,
    labels = c("Q1", "Q2", "Q3", "Q4"),
    include.lowest = TRUE
  )

  # Merge expression data with IC50 data (averaged)
  merged_df <- merge(expression_df, ic50_drugs_gene_list, by.x = "Patient", by.y = "subject_id")

  # Perform Kruskal-Wallis and Pairwise Wilcoxon Tests for Each Drug
  pairwise_results <- data.frame()

  # Pivot to long format for easier manipulation
  drug_data_long <- merged_df %>%
    pivot_longer(cols = starts_with("CHZ"):Vorinostat, names_to = "Drug", values_to = "LC50")

  # Loop over each drug to perform Kruskal-Wallis and pairwise Wilcoxon tests
  for (drug in unique(drug_data_long$Drug)) {
    drug_data <- drug_data_long %>% filter(Drug == drug)
    kw_test <- kruskal_test(data = drug_data, LC50 ~ Quartile)
    
    if (kw_test$p < 0.05) {
      pw_test <- drug_data %>%
        pairwise_wilcox_test(LC50 ~ Quartile, p.adjust.method = "fdr") %>%
        mutate(Drug = drug)
      pairwise_results <- bind_rows(pairwise_results, pw_test)
    }
  }

  # Store the results for the current gene
  all_pairwise_results[[gene]] <- pairwise_results

  # Check if pairwise_results is populated correctly
  if (nrow(pairwise_results) == 0) {
    message(paste("No significant pairwise comparisons for gene", gene))
  }
}

# Now proceed to plotting for genes with significant pairwise comparisons
for (gene in target_genes) {
  pairwise_results <- all_pairwise_results[[gene]]
  
  # Check if there are significant results to plot
  if (is.null(pairwise_results) || nrow(pairwise_results) == 0) {
    next
  }
  
  pairwise_results <- pairwise_results %>%
    dplyr::select(Drug, group1, group2, p, p.adj, p.adj.signif) %>%
    arrange(Drug, p.adj)
  
  # Extract unique drugs with significant pairwise comparisons
  significant_drugs <- unique(pairwise_results$Drug[pairwise_results$p.adj.signif != "ns"])
  
  # Define the base directory
  base_dir <- "/Users/cojulian/Desktop/Tzu_projects/Matt/Integrated_D_R_curtis/tasks_results/task6_stjudes_pharmacotypes/avarage_imputs_all_quartiles"

  # Create a directory for the gene within the base directory
  dir.create(file.path(base_dir, gene), showWarnings = FALSE)
  
  # Loop over each significant drug to create and display plots
  for (drug in significant_drugs) {
    drug_data <- merged_df %>%
      dplyr::select(Patient, Quartile, !!sym(drug)) %>%
      rename(LC50 = !!sym(drug))
    
    n_data <- drug_data %>%
      group_by(Quartile) %>%
      summarise(n = n())
    
    comparison_data <- pairwise_results %>%
      filter(Drug == drug) %>%
      mutate(p.adj.signif = ifelse(p.adj.signif == "NS", "ns", p.adj.signif),  # Ensure lowercase "ns"
             y.position = max(drug_data$LC50, na.rm = TRUE) * 1.1)
    
    comparisons <- comparison_data %>%
      dplyr::select(group1, group2) %>%
      split(seq(nrow(comparison_data))) %>%
      lapply(unlist)
    
    plot <- ggplot(drug_data, aes(x = Quartile, y = LC50, fill = Quartile)) +
      geom_boxplot(outlier.shape = NA) +
      viridis::scale_fill_viridis(discrete = TRUE, option = "D") +
      labs(
        title = paste("LC50 Distribution for", drug, "Across Quartiles of", gene, "expression (Averaged)"),
        x = "Quartile",
        y = "LC50"
      ) +
      theme_minimal() +
      theme(
        plot.title = element_text(hjust = 0.5, size = 14),
        axis.text = element_text(size = 12),
        axis.title = element_text(size = 13)
      ) +
      geom_text(
        data = n_data,
        aes(x = Quartile, y = min(drug_data$LC50, na.rm = TRUE) * 0.95, label = paste0("n = ", n)),
        color = "black",
        size = 4
      ) +
      geom_signif(
        comparisons = comparisons,
        annotations = comparison_data$p.adj.signif,  # Explicitly use the modified lowercase "ns" column
        y_position = seq(
          from = max(drug_data$LC50, na.rm = TRUE) * 1.05, 
          by = 0.08, 
          length.out = nrow(comparison_data)
        ),  # Space out annotations
        map_signif_level = TRUE,
        tip_length = 0.005,
        textsize = 5
      )
    
    print(plot)
    
    # Save the plot to a file in the respective gene directory
    ggsave(
  filename = file.path(base_dir, gene, paste0("LC50_Distribution_", drug, "_", gene, "_Averaged.pdf")), 
  plot = plot, 
  width = 8,  # Adjust the width as needed
  height = 5, # Adjust the height as needed
  units = "in" # Use inches for measurement
)
  }
}

```
## All genes across all drugs NON-AVERAGE

```{r}
# Define the list of target genes
target_genes <- c("GEM", "INHBE", "RASL11A", "CYP7B1", "NFKBIE", 
                  "ZBTB10", "CTSB", "OLAH", "FAM76B")

# Initialize a list to store pairwise results for each gene
all_pairwise_results <- list()

# Loop through each gene to perform statistical tests
for (gene in target_genes) {
  # Extract and process expression data for the current gene
  if (gene %in% rnaseq_drugs_list$GeneName) {
    gene_expression <- rnaseq_drugs_list[rnaseq_drugs_list$GeneName == gene, ]
    expression_values <- gene_expression[, !(names(gene_expression) %in% c("GeneID", "GeneName")), drop = FALSE]
    expression_df <- as.data.frame(t(expression_values))
    colnames(expression_df) <- "Expression"
    expression_df$Patient <- rownames(expression_df)
  } else {
    stop(paste("Gene", gene, "not found in the dataset."))
  }

  # Define quartiles for the current gene expression
  quantiles <- quantile(expression_df$Expression, probs = seq(0, 1, 0.25), na.rm = TRUE)
  if (length(unique(quantiles)) < 5) {
    message(paste("Quantiles are not unique for gene", gene, "- skipping."))
    next
  }
  
  expression_df$Quartile <- cut(
    expression_df$Expression,
    breaks = quantiles,
    labels = c("Q1", "Q2", "Q3", "Q4"),
    include.lowest = TRUE
  )

  # Merge expression data with IC50 data
  merged_df <- merge(expression_df, ic50_drugs_gene, by.x = "Patient", by.y = "subject_id")

  # Perform Kruskal-Wallis and Pairwise Wilcoxon Tests for Each Drug
  pairwise_results <- data.frame()

  # Pivot to long format for easier manipulation
  drug_data_long <- merged_df %>%
    pivot_longer(cols = starts_with("CHZ"):Vorinostat, names_to = "Drug", values_to = "LC50")

  # Loop over each drug to perform Kruskal-Wallis and pairwise Wilcoxon tests
  for (drug in unique(drug_data_long$Drug)) {
    drug_data <- drug_data_long %>% filter(Drug == drug)
    kw_test <- kruskal_test(data = drug_data, LC50 ~ Quartile)
    
    if (kw_test$p < 0.05) {
      pw_test <- drug_data %>%
        pairwise_wilcox_test(LC50 ~ Quartile, p.adjust.method = "fdr") %>%
        mutate(Drug = drug)
      pairwise_results <- bind_rows(pairwise_results, pw_test)
    }
  }

  # Store the results for the current gene
  all_pairwise_results[[gene]] <- pairwise_results

  # Check if pairwise_results is populated correctly
  if (nrow(pairwise_results) == 0) {
    message(paste("No significant pairwise comparisons for gene", gene))
  }
}

# Now proceed to plotting for genes with significant pairwise comparisons
for (gene in target_genes) {
  pairwise_results <- all_pairwise_results[[gene]]
  
  # Check if there are significant results to plot
  if (is.null(pairwise_results) || nrow(pairwise_results) == 0) {
    next
  }
  
  pairwise_results <- pairwise_results %>%
    dplyr::select(Drug, group1, group2, p, p.adj, p.adj.signif) %>%
    arrange(Drug, p.adj)
  
  # Extract unique drugs with significant pairwise comparisons
  significant_drugs <- unique(pairwise_results$Drug[pairwise_results$p.adj.signif != "ns"])
  
  # Define the base directory
  base_dir <- "/Users/cojulian/Desktop/Tzu_projects/Matt/Integrated_D_R_curtis/tasks_results/task6_stjudes_pharmacotypes/noavarage_imputs_all_quartiles"

  # Create a directory for the gene within the base directory
  dir.create(file.path(base_dir, gene), showWarnings = FALSE)
  
  # Loop over each significant drug to create and display plots
  for (drug in significant_drugs) {
    drug_data <- merged_df %>%
      dplyr::select(Patient, Quartile, !!sym(drug)) %>%
      rename(LC50 = !!sym(drug))
    
    n_data <- drug_data %>%
      group_by(Quartile) %>%
      summarise(n = n())
    
    comparison_data <- pairwise_results %>%
      filter(Drug == drug) %>%
      mutate(p.adj.signif = ifelse(p.adj.signif == "NS", "ns", p.adj.signif),  # Ensure lowercase "ns"
             y.position = max(drug_data$LC50, na.rm = TRUE) * 1.1)
    
    comparisons <- comparison_data %>%
      dplyr::select(group1, group2) %>%
      split(seq(nrow(comparison_data))) %>%
      lapply(unlist)
    
    plot <- ggplot(drug_data, aes(x = Quartile, y = LC50, fill = Quartile)) +
      geom_boxplot(outlier.shape = NA) +
      viridis::scale_fill_viridis(discrete = TRUE, option = "D") +
      labs(
        title = paste("LC50 Distribution for", drug, "Across Quartiles of", gene, "expression (Non-Averaged)"),
        x = "Quartile",
        y = "LC50"
      ) +
      theme_minimal() +
      theme(
        plot.title = element_text(hjust = 0.5, size = 14),
        axis.text = element_text(size = 12),
        axis.title = element_text(size = 13)
      ) +
      geom_text(
        data = n_data,
        aes(x = Quartile, y = min(drug_data$LC50, na.rm = TRUE) * 0.95, label = paste0("n = ", n)),
        color = "black",
        size = 4
      ) +
      geom_signif(
        comparisons = comparisons,
        annotations = comparison_data$p.adj.signif,  # Explicitly use the modified lowercase "ns" column
        y_position = seq(
          from = max(drug_data$LC50, na.rm = TRUE) * 1.05, 
          by = 0.08, 
          length.out = nrow(comparison_data)
        ),  # Space out annotations
        map_signif_level = TRUE,
        tip_length = 0.005,
        textsize = 5
      )
    
    print(plot)
    
    # Save the plot to a file in the respective gene directory
    ggsave(
      filename = file.path(base_dir, gene, paste0("LC50_Distribution_", drug, "_", gene, "_NonAveraged.pdf")), 
      plot = plot, 
      width = 8,  # Adjust the width as needed
      height = 5, # Adjust the height as needed
      units = "in" # Use inches for measurement
    )
  }
}
```


## AVg and non Avg-all of the genes for Q1 and Q4

```{r}
# Define the list of target genes
target_genes <- c("GEM", "INHBE", "RASL11A", "CYP7B1", "NFKBIE", 
                  "ZBTB10", "CTSB", "OLAH", "FAM76B")

# Define the base directory for saving plots
base_dir <- "/Users/cojulian/Desktop/Tzu_projects/Matt/Integrated_D_R_curtis/tasks_results/task6_stjudes_pharmacotypes"

# Function to generate and save plots
generate_plots <- function(data, pairwise_results, base_dir, sub_dir, data_type) {
  # Check if `pairwise_results` is empty
  if (nrow(pairwise_results) == 0) {
    stop("The `pairwise_results` dataframe is empty. Ensure statistical analysis is complete before running this script.")
  }
  
  # Loop through each gene in target_genes
  for (target_gene in target_genes) {
    
    # Check if the target gene exists in the data
    if (target_gene %in% data$GeneName) {
      # Extract expression data for the target gene
      gene_expression <- data[data$GeneName == target_gene, ]
      expression_values <- gene_expression[, !(names(gene_expression) %in% c("GeneID", "GeneName")), drop = FALSE]
      expression_df <- as.data.frame(t(expression_values))
      colnames(expression_df) <- "Expression"
      expression_df$Patient <- rownames(expression_df)
      
      # Define quartiles for the current gene expression
      quantiles <- quantile(expression_df$Expression, probs = seq(0, 1, 0.25), na.rm = TRUE)
      if (length(unique(quantiles)) < 5) {
        message(paste("Quantiles are not unique for gene", target_gene, "- skipping."))
        next
      }
      
      expression_df$Quartile <- cut(
        expression_df$Expression,
        breaks = quantiles,
        labels = c("Q1", "Q2", "Q3", "Q4"),
        include.lowest = TRUE
      )
      
      # Filter to only Q1 and Q4
      expression_df <- expression_df %>% filter(Quartile %in% c("Q1", "Q4"))
      
      # Merge with LC50 data
      merged_df <- merge(expression_df, ic50_drugs_gene, by.x = "Patient", by.y = "subject_id")
      
      # Filter `pairwise_results` for drugs with significant Q1 vs Q4 comparisons
      significant_drugs <- pairwise_results %>%
        filter(group1 == "Q1", group2 == "Q4") %>%
        dplyr::select(Drug, p.adj.signif) %>%
        distinct()
  
      # Check if there are significant drugs
      if (nrow(significant_drugs) == 0) {
        message(paste("No significant Q1 vs Q4 comparisons found for gene", target_gene))
        next
      }
  
      # Create a subdirectory for the target gene in the base directory
      gene_dir <- file.path(base_dir, sub_dir, target_gene)
      dir.create(gene_dir, showWarnings = FALSE, recursive = TRUE)
      
      # Loop through each drug with a significant Q1 vs Q4 comparison
      for (drug_info in unique(significant_drugs$Drug)) {
        drug <- drug_info  # Extract drug name
  
        # Extract relevant columns and calculate sample size
        drug_data <- merged_df %>%
         dplyr::select(Patient, Quartile, !!sym(drug)) %>%
          rename(LC50 = !!sym(drug))
        
        # Ensure there is data for both Q1 and Q4
        if (nrow(drug_data) < 2) {
          message(paste("Insufficient data for Q1 and Q4 for drug", drug, "- skipping."))
          next
        }
  
        n_data <- drug_data %>%
          group_by(Quartile) %>%
          summarise(n = n())
        
        # Fetch the Q1 vs Q4 comparison for this drug
        comparison_data <- significant_drugs %>%
          filter(Drug == drug) %>%
          mutate(y.position = max(drug_data$LC50, na.rm = TRUE) * 1.1)
        
        # Define the comparison for geom_signif
        comparisons <- list(c("Q1", "Q4"))
  
        # Generate and display the plot
        plot <- ggplot(drug_data, aes(x = Quartile, y = LC50, fill = Quartile)) +
          geom_boxplot(outlier.shape = NA) +
          viridis::scale_fill_viridis(discrete = TRUE, option = "D") +
          labs(
            title = paste("LC50 Distribution for", drug, "Between Q1 and Q4 of", target_gene, "expression (", data_type, ")"),
            x = "Quartile",
            y = "LC50"
          ) +
          theme_minimal() +
          theme(
            plot.title = element_text(hjust = 0.5, size = 14),
            axis.text = element_text(size = 12),
            axis.title = element_text(size = 13)
          ) +
          geom_text(
            data = n_data,
            aes(x = Quartile, y = min(drug_data$LC50, na.rm = TRUE) * 0.95, label = paste0("n = ", n)),
            color = "black",
            size = 4
          ) +
          geom_signif(
            comparisons = comparisons,
            annotations = comparison_data$p.adj.signif,
            y_position = comparison_data$y.position,
            map_signif_level = TRUE,
            tip_length = 0.005,
            textsize = 5
          )
        
        # Display the plot
        print(plot)
        
        # Save the plot to a file in the respective gene directory
        plot_filename <- file.path(gene_dir, paste0("LC50_Distribution_", drug, "_", target_gene, "_", data_type, ".pdf"))
        ggsave(
          filename = plot_filename, 
          plot = plot, 
          width = 8,  # Adjust the width as needed
          height = 5, # Adjust the height as needed
          units = "in" # Use inches for measurement
        )
      }
    } else {
      message(paste("Gene", target_gene, "not found in the dataset."))
    }
  }
}

# Generate plots for averaged data
generate_plots(rnaseq_drugs_list, pairwise_results, base_dir, "avarage_imputs_q1_q4", "averaged")

# Generate plots for non-averaged data
generate_plots(rnaseq_drugs_list, pairwise_results, base_dir, "noavarage_imputs_q1_q4", "non-averaged")

```




# Task 7. November 18

## Task 7.1 "Using the low hypodiploid samples from St Jude FKPM data, can you split that group into GEM expression quartiles. Can you then run Hallmarks pathway analysis on each quartile."

```{r}
# Loading the huge dataframe again
stjudes_df
```

```{r}
# Keeping only LH samples

# Filter columns that contain "Low hyperdiploid"
LH_only_sjudes_df <- stjudes_df %>%
 dplyr::select(Gene, contains("Low hyperdiploid"))
```


```{r}

# Step 1: Extract the GEM row from the Low hyperdiploid dataframe
gem_row <- LH_only_sjudes_df %>%
  filter(Gene == "GEM")

# Step 2: Transpose the GEM expression values to a vector for quartile calculation
gem_values <- as.numeric(unlist(gem_row %>%dplyr::select(-Gene)))

# Step 3: Calculate quartiles and assign each value to its quartile group
gem_quartiles <- cut(gem_values, 
                     breaks = quantile(gem_values, probs = seq(0, 1, 0.25), na.rm = TRUE), 
                     include.lowest = TRUE, 
                     labels = c("Q1", "Q2", "Q3", "Q4"))

# Step 4: Combine GEM values with quartile assignments
gem_quartile_df <- data.frame(
  Sample = names(gem_row)[-1],  # Column names as sample IDs
  Expression = gem_values,
  Quartile = gem_quartiles
)

# Display the quartile assignments
print(gem_quartile_df)

```

```{r}
# plotting that
ggplot(gem_quartile_df, aes(x = Quartile, y = Expression)) +
  geom_boxplot() +
  labs(title = "GEM Expression by Quartile in LH samples",
       x = "Quartile",
       y = "Expression")
```

```{r}
gene_quartile_summary <- LH_only_sjudes_df %>%
  pivot_longer(cols = -Gene, names_to = "Sample", values_to = "Expression") %>%
  left_join(gem_quartile_df, by = "Sample") %>%
  group_by(Gene, Quartile) %>%
  summarize(Average_Expression = mean(Expression.x, na.rm = TRUE), .groups = 'drop')

```

## Testing for Q1 first:

```{r, fig.width=8}
q1_specific_genes <- gene_quartile_summary %>%
  filter(Quartile == "Q1") %>%
  arrange(desc(Average_Expression))


# Step 1: Prepare Ranked Gene List
# Filter and rank genes for Q1
q1_genes <- q1_specific_genes %>%
 dplyr::select(Gene, Average_Expression) %>%
  arrange(desc(Average_Expression))

# Create a named vector of ranked genes
gene_ranks <- q1_genes %>%
  deframe()

# Break ties in ranked genes by adding small random noise
set.seed(123)  # For reproducibility
gene_ranks <- gene_ranks + rnorm(length(gene_ranks), mean = 0, sd = 1e-5)

# Step 2: Retrieve Hallmark Pathways
# Retrieve Hallmark gene sets for Homo sapiens
hallmark_pathways <- msigdbr(species = "Homo sapiens", category = "H")

# Format pathways for fgsea
hallmark_pathway_list <- hallmark_pathways %>%
  split(x = .$gene_symbol, f = .$gs_name)

# Step 3: Perform GSEA
# Run fgseaMultilevel for better precision
fgsea_results <- fgsea(
  pathways = hallmark_pathway_list,
  stats = gene_ranks,
  minSize = 15,
  maxSize = 500,
  eps = 0  # Increases precision for very small p-values
)

# Step 4: Process Results
# Convert results to a tidy tibble and filter significant pathways
fgsea_tidy <- fgsea_results %>%
  as_tibble() %>%
  arrange(padj)

# Filter significant pathways (padj < 0.05)
significant_pathways <- fgsea_tidy %>%
  filter(padj < 0.05)

# Step 5: Visualize Top Enriched Pathways
# Bar plot of top 20 significant pathways by NES
fgsea_tidy %>%
  filter(padj < 0.05) %>%          # Filter significant pathways
  slice_min(padj, n = 20) %>%     # Select top 20 pathways by adjusted p-value
  ggplot(aes(x = reorder(pathway, NES), y = NES, fill = NES > 0)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top Enriched Hallmark Pathways of Q1 GEM expression",
    x = "Pathway",
    y = "Normalized Enrichment Score (NES)"
  ) +
  scale_fill_manual(values = c("TRUE" = "#61BB6F", "FALSE" = "#EC7012")) +
  theme_minimal()

```


## Trying a loop for all quartiles
```{r}
# Retrieve Hallmark gene sets for Homo sapiens
hallmark_sets <- msigdbr(species = "Homo sapiens", category = "H")

# Format the gene sets for fgsea
fgsea_hallmark_sets <- hallmark_sets %>%
  split(x = .$gene_symbol, f = .$gs_name)

# Function to perform GSEA for a given quartile
perform_gsea <- function(quartile) {
  # Filter and rank genes for the given quartile
  quartile_genes <- gene_quartile_summary %>%
    filter(Quartile == quartile) %>%
    dplyr::select(Gene, Average_Expression) %>%
    arrange(desc(Average_Expression))
  
  # Create a named vector of ranked genes
  gene_ranks <- quartile_genes %>%
    deframe()
  
  # Break ties in ranked genes by adding small random noise
  set.seed(123)  # For reproducibility
  gene_ranks <- gene_ranks + runif(length(gene_ranks), min = -1e-6, max = 1e-6)
  
  # Perform GSEA with eps set to zero
  fgsea_results <- fgsea(pathways = fgsea_hallmark_sets, stats = gene_ranks, eps = 0)
  
# Tidy the results
fgsea_results_tidy <- fgsea_results %>%
  as_tibble() %>%
  arrange(padj)

# Extract leading edge genes and include their expression for the specific quartile
leading_edge_genes <- fgsea_results %>%
  filter(padj < 0.05) %>%
  dplyr::select(pathway, leadingEdge) %>%
  unnest(cols = c(leadingEdge)) %>%
  distinct() %>%
  left_join(gene_quartile_summary %>% filter(Quartile == quartile) %>% distinct(Gene, .keep_all = TRUE), by = c("leadingEdge" = "Gene"))
  
  # Save the plot
  output_dir <- "/Users/cojulian/Desktop/Tzu_projects/Matt/Integrated_D_R_curtis/tasks_results/task7/task7.1"
  # Save leading edge genes to CSV
  write_csv(leading_edge_genes, paste0(output_dir, "/Leading_Edge_Genes_", quartile, ".csv"))
  
  # Plot the top 20 pathways
  plot <- fgsea_results_tidy %>%
    filter(pval < 0.05) %>%
    arrange(padj) %>%
    head(n = 20) %>%
    ggplot(aes(reorder(pathway, NES), NES)) +
    geom_col(aes(fill = NES > 0)) +
    scale_fill_manual(values = c("TRUE" = "#61BB6F", "FALSE" = "#EC7012")) +
    coord_flip() +
    labs(x = "Pathway", y = "Normalized Enrichment Score", title = paste("Top 20 Hallmark Pathways of", quartile, "GEM expression with padj<0.05")) +
    theme_minimal()
  
  
  ggsave(filename = paste0(output_dir, "/Top_20_Hallmark_Pathways_", quartile, ".pdf"), plot = plot, width = 16, height = 9)
}

# Perform GSEA for each quartile
quartiles <- c("Q1", "Q2", "Q3", "Q4")
lapply(quartiles, perform_gsea)
```
### Task 7.1.1 perform DE analysis of Q1 versus Q4 of Task 7.1
```{r}
# Create metadata for Q1 and Q4 samples
meta_data_q1_q4 <- gem_quartile_df %>%
  filter(Quartile %in% c("Q1", "Q4")) %>%
  mutate(condition = ifelse(Quartile == "Q1", "Q1", "Q4"))

# Ensure the sample names match the count data
meta_data_q1_q4 <- meta_data_q1_q4 %>%
  rename(sample = Sample)

# Filter the count data to include only Q1 and Q4 samples
count_data_q1_q4 <- LH_only_sjudes_df %>%
 dplyr::select(Gene, all_of(meta_data_q1_q4$sample))

# Create a design matrix
design <- model.matrix(~ 0 + meta_data_q1_q4$condition)
colnames(design) <- c("Q1", "Q4")

# Ensure the row names of meta_data_q1_q4 match the column names of count_data_q1_q4
rownames(meta_data_q1_q4) <- meta_data_q1_q4$sample
count_data_q1_q4 <- count_data_q1_q4 %>%
  column_to_rownames(var = "Gene")
```

```{r}
# Fit the linear model
fit <- lmFit(count_data_q1_q4, design)

# Create a contrast matrix
contrast_matrix <- makeContrasts(Q4 - Q1, levels = design)

# Fit the contrasts
fit2 <- contrasts.fit(fit, contrast_matrix)
fit2 <- eBayes(fit2)
```

```{r}
results <- topTable(fit2, adjust = "fdr", number = Inf)
summary(results)

# Convert results to a data frame
de_results_q1_vs_q4 <- as.data.frame(results)
```
### Volcano plot

```{r, fig.width=18}
# Define custom colors, sizes, and alphas for the plot
cols <- c("up" = "#ffad73", "down" = "#26b3ff", "ns" = "grey")
sizes <- c("up" = 2, "down" = 2, "ns" = 1)
alphas <- c("up" = 1, "down" = 1, "ns" = 0.5)

# Prepare the data for plotting
volcano_data <- de_results_q1_vs_q4 %>%
  rownames_to_column("GeneSymbol") %>%
  dplyr::select(GeneSymbol, logFC, adj.P.Val) %>%
  mutate(FoldChange = 2^logFC) %>%
  mutate(GeneType = case_when(
    logFC >= 1 & adj.P.Val <= 0.05 ~ "up",
    logFC <= -1 & adj.P.Val <= 0.05 ~ "down",
    TRUE ~ "ns"
  ))

# Create the volcano plot
volcano_plot <- volcano_data %>%
  ggplot(aes(x = logFC, y = -log10(adj.P.Val),
             fill = GeneType, size = GeneType, alpha = GeneType)) +
  geom_point(shape = 21, colour = "black") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
  scale_fill_manual(values = cols) +
  scale_size_manual(values = sizes) +
  scale_alpha_manual(values = alphas) +
  geom_text_repel(
    data = filter(volcano_data, adj.P.Val < 0.05),
    aes(x = logFC, y = -log10(adj.P.Val), label = GeneSymbol),
    size = 5,
    color = "darkblue",
    max.overlaps = 10,
    min.segment.length = Inf,
    inherit.aes = FALSE
  ) +
  labs(title = "Volcano Plot of DE Genes: Q1 vs Q4 with adj.P.Val <= 0.05 and logFC > |1|",
       x = "log2(Fold Change)",
       y = "-log10(Adjusted P-value)") +
  theme_minimal(base_family = "Arial") +
  theme(
    panel.background = element_rect(fill = "white", colour = NA),
    plot.background = element_rect(fill = "white", colour = NA),
    legend.position = "bottom",
    plot.title = element_text(size = 20, face = "bold"),
    axis.title.x = element_text(size = 18),
    axis.title.y = element_text(size = 18),
    axis.text = element_text(size = 14),
    legend.text = element_text(size = 14),
    legend.title = element_text(size = 16)
  )

# Display the plot
print(volcano_plot)

# Save the plot with high resolution and proper background settings
ggsave("volcano_plot_q1_vs_q4.png", plot = volcano_plot, width = 16, height = 9, dpi = 300, bg = "white")
```

### Now hallmakrs on those DEGs

```{r}
# Retrieve Hallmark gene sets for Homo sapiens
hallmark_pathways <- msigdbr(species = "Homo sapiens", category = "H")

# Format pathways for fgsea
hallmark_pathway_list <- hallmark_pathways %>%
  split(x = .$gene_symbol, f = .$gs_name)

filtered_genes <- de_results_q1_vs_q4 %>%
  filter(abs(logFC) > 1) %>%
  rownames_to_column("GeneSymbol")
```

### Prepare gene ranks

```{r}
# Rank genes based on log2 fold change
gene_ranks <- filtered_genes %>%
  dplyr::select(GeneSymbol, logFC) %>%
  arrange(desc(logFC)) %>%
  deframe()

# Ensure all values are finite
gene_ranks <- gene_ranks[is.finite(gene_ranks)]

# Add small random noise to break ties
set.seed(123)
gene_ranks <- gene_ranks + runif(length(gene_ranks), min = -1e-6, max = 1e-6)
```

```{r}
# Perform GSEA with adjusted parameters
fgsea_results <- fgsea(
  pathways = hallmark_pathway_list,
  stats = gene_ranks,
  minSize = 10,  # Reduced minSize to include smaller pathways
  maxSize = 1000,  # Increased maxSize to include larger pathways
  eps = 0
)

# Tidy the results
fgsea_tidy_7.1.1 <- fgsea_results %>%
  as_tibble() %>%
  arrange(padj)

 
```


```{r, fig.width=10}
# Plot the top 20 pathways
plot_7.1.1 <- fgsea_tidy_7.1.1 %>%
  slice_min(padj, n = 20) %>%
  ggplot(aes(x = reorder(pathway, NES), y = NES, fill = NES > 0)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top Enriched Hallmark Pathways: Q1 vs Q4 (logFC > 1)",
    x = "Pathway",
    y = "Normalized Enrichment Score (NES)"
  ) +
  scale_fill_manual(values = c("TRUE" = "#61BB6F", "FALSE" = "#EC7012")) +
  theme_minimal()

# Display the plot
print(plot_7.1.1)

# Save the plot with high resolution and proper background settings
#ggsave("hallmark_pathways_q1_vs_q4.png", plot = plot, width = 16, height = 9, dpi = 300, bg = "white")
```


## Task 7.2  Run Hallmarks pathway analysis on diagnosis versus relapse leukemia from the individual scRNA-seq LH samples (LH001, LH002, LH003).

```{r}
matts_df_LH_7.2 <- as.data.frame(filtered_bulk_matrix) %>% 
 dplyr::select(starts_with("LH"))
```

### Subsetting Metadata
```{r}
# Create metadata (e.g., Diagnosis and Relapse labels for your columns)
meta_data_7.2 <- data.frame(
  sample = colnames(matts_df_LH_7.2)[-1],  # Exclude the Gene column
  condition = ifelse(grepl("-D", colnames(matts_df_LH_7.2)[-1]), "Diagnosis", "Relapse"),
  stringsAsFactors = FALSE
)
```
### Running DESeq   
```{r}
# Create DESeq2 dataset
dds_7.2 <- DESeqDataSetFromMatrix(countData = matts_df_LH_7.2[,-1], colData = meta_data_7.2, design = ~ condition)

# Run DESeq2
dds_7.2 <- DESeq(dds_7.2)

# Get DE results
res_7.2 <- results(dds_7.2)
summary(res_7.2)

de_results_all_D_vs_R_7.2 <- as.data.frame(res_7.2)
```

```{r, fig.width=10}
# Retrieve Hallmark gene sets for Homo sapiens
hallmark_pathways_7.2 <- msigdbr(species = "Homo sapiens", category = "H")

# Format pathways for fgsea
hallmark_pathway_list_7.2 <- hallmark_pathways_7.2 %>%
  split(x = .$gene_symbol, f = .$gs_name)

# Rank genes based on log2 fold change
gene_ranks_7.2 <- de_results_all_D_vs_R_7.2 %>%
  rownames_to_column("GeneSymbol") %>%
 dplyr::select(GeneSymbol, log2FoldChange) %>%
  arrange(desc(log2FoldChange)) %>%
  deframe()

# Ensure all values are finite
gene_ranks_7.2 <- gene_ranks_7.2[is.finite(gene_ranks_7.2)]

# Add small random noise to break ties
set.seed(123)
gene_ranks_7.2 <- gene_ranks_7.2 + runif(length(gene_ranks_7.2), min = -1e-6, max = 1e-6)

# Perform GSEA with adjusted parameters
fgsea_results_7.2 <- fgsea(
  pathways = hallmark_pathway_list_7.2,
  stats = gene_ranks_7.2,
  minSize = 10,  # Reduced minSize to include smaller pathways
  maxSize = 1000,  # Increased maxSize to include larger pathways
  eps = 0
)
# Tidy the results
fgsea_tidy_7.2 <- fgsea_results_7.2 %>%
  as_tibble() %>%
  arrange(padj)

# Filter significant pathways (padj < 0.05)
significant_pathways_7.2 <- fgsea_tidy_7.2 %>%
  filter(padj < 0.1)

# Plot the top 20 pathways
plot_7.2 <- fgsea_tidy_7.2 %>%
  filter(padj < 0.1) %>%
  slice_min(padj, n = 20) %>%
  ggplot(aes(x = reorder(pathway, NES), y = NES, fill = NES > 0)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top Enriched Hallmark Pathways: Diagnosis vs Relapse (padj < 0.1)",
    x = "Pathway",
    y = "Normalized Enrichment Score (NES)"
  ) +
  scale_fill_manual(values = c("TRUE" = "#61BB6F", "FALSE" = "#EC7012")) +
  theme_minimal()

plot_7.2

# Save the plot
#output_dir_7.2 <- "/Users/cojulian/Desktop/Tzu_projects/Matt/Integrated_D_R_curtis/tasks_results/task7/task7.1"
#ggsave(filename = paste0(output_dir_7.2, "/Top_20_Hallmark_Pathways_Diagnosis_vs_Relapse.pdf"), plot = plot_7.2)
```

### Task 7.2.1 Run each sample separately

 
```{r}

# Get list of pairs (Dx and Rx)
pair_ids <- unique(gsub("-[DR]", "", colnames(matts_df_LH_7.2)))
pairs <- lapply(pair_ids, function(id) {
  list(
    Dx = paste0(id, "-D"),
    Rx = paste0(id, "-R")
  )
})
```

```{r}
hallmark_pathways_7.2 <- msigdbr(species = "Homo sapiens", category = "H")

# Format pathways for fgsea
hallmark_pathway_list_7.2 <- hallmark_pathways_7.2 %>%
  split(x = .$gene_symbol, f = .$gs_name)

# Step 3: Loop Through Each Pair and Perform GSEA
results_list <- list()
plots_list <- list()

for (pair in pairs) {
  dx_col <- pair$Dx
  rx_col <- pair$Rx
  
  # Ensure both columns exist in the dataset
  if (dx_col %in% colnames(matts_df_LH_7.2) && rx_col %in% colnames(matts_df_LH_7.2)) {
    # Calculate log2 fold changes
    gene_ranks <- log2(matts_df_LH_7.2[[rx_col]] + 1) - log2(matts_df_LH_7.2[[dx_col]] + 1)
    names(gene_ranks) <- rownames(matts_df_LH_7.2)  # Assign gene names
    
    # Add small random noise to break ties
    set.seed(123)
    gene_ranks <- gene_ranks + runif(length(gene_ranks), min = -1e-6, max = 1e-6)
    
    # Perform GSEA
    fgsea_results <- fgsea(
      pathways = hallmark_pathway_list_7.2,
      stats = gene_ranks,
      minSize = 10,
      maxSize = 1000,
      eps = 0
    )
    
    # Tidy results
    fgsea_tidy <- fgsea_results %>%
      as_tibble() %>%
      arrange(padj)
    
    # Filter significant pathways (padj < 0.1)
    significant_pathways <- fgsea_tidy %>%
      filter(padj < 0.05)
    
    # Save results for this pair
    results_list[[paste0(dx_col, "_vs_", rx_col)]] <- fgsea_tidy
    
    # Plot top 20 pathways
    plot <- fgsea_tidy %>%
      filter(padj < 0.1) %>%
      slice_min(padj, n = 20) %>%
      ggplot(aes(x = reorder(pathway, NES), y = NES, fill = NES > 0)) +
      geom_col() +
      coord_flip() +
      labs(
        title = paste0("Top Enriched Pathways: ", dx_col, " vs ", rx_col),
        x = "Pathway",
        y = "Normalized Enrichment Score (NES)"
      ) +
      scale_fill_manual(values = c("TRUE" = "#61BB6F", "FALSE" = "#EC7012")) +
      theme_minimal()
    
    plots_list[[paste0(dx_col, "_vs_", rx_col)]] <- plot
  } else {
    warning(paste("Columns not found for pair:", dx_col, rx_col))
  }
}

```


### Task 7.2.2 Take avarage expression and make hallmakrs

```{r}
# Separate D and R samples
d_samples <- matts_df_LH_7.2 %>%
  rownames_to_column("Gene") %>% 
  dplyr::select(Gene, contains("-D"))

r_samples <- matts_df_LH_7.2 %>%
  rownames_to_column("Gene") %>% 
  dplyr::select(Gene, contains("-R"))

# Calculate average expression for D samples
d_avg_expression <- d_samples %>%
  rowwise() %>%
  mutate(Average_D = mean(c_across(starts_with("LH")), na.rm = TRUE)) %>%
  dplyr::select(Gene, Average_D)

# Calculate average expression for R samples
r_avg_expression <- r_samples %>%
  rowwise() %>%
  mutate(Average_R = mean(c_across(starts_with("LH")), na.rm = TRUE)) %>%
  dplyr::select(Gene, Average_R)

# Combine the average expressions into one dataframe
avg_expression_df <- d_avg_expression %>%
  inner_join(r_avg_expression, by = "Gene")

# Calculate log2 fold change
avg_expression_df <- avg_expression_df %>%
  mutate(log2FC = log2(Average_R / Average_D))

# Display the dataframe with log2FC
print(avg_expression_df)

# Filter genes based on log2FC > |1|
filtered_avg_expression_df <- avg_expression_df %>%
  filter(abs(log2FC) > 1)

# Display the dataframe with log2FC
print(filtered_avg_expression_df)
```

#### Running the GSEA

```{r, fig.width=15}
# Retrieve Hallmark gene sets for Homo sapiens
hallmark_pathways_7.2.2 <- msigdbr(species = "Homo sapiens", category = "H")

# Format pathways for fgsea
hallmark_pathway_list_7.2.2 <- hallmark_pathways_7.2.2 %>%
  split(x = .$gene_symbol, f = .$gs_name)

# Rank genes based on log2 fold change
gene_ranks_7.2.2 <- avg_expression_df %>%
  dplyr::select(Gene, log2FC) %>%
  arrange(desc(log2FC)) %>%
  deframe()

# Ensure all values are finite
gene_ranks_7.2.2 <- gene_ranks_7.2.2[is.finite(gene_ranks_7.2.2)]

# Add small random noise to break ties
set.seed(123)
gene_ranks_7.2.2 <- gene_ranks_7.2.2 + runif(length(gene_ranks_7.2.2), min = -1e-6, max = 1e-6)

# Perform GSEA with adjusted parameters
fgsea_results_7.2.2 <- fgsea(
  pathways = hallmark_pathway_list_7.2.2,
  stats = gene_ranks_7.2.2,
  minSize = 10,  # Reduced minSize to include smaller pathways
  maxSize = 1000,  # Increased maxSize to include larger pathways
  eps = 0
)

# Tidy the results
fgsea_tidy_7.2.2 <- fgsea_results_7.2.2 %>%
  as_tibble() %>%
  arrange(padj)

# Plot the top pathways
plot_7.2.2 <- fgsea_tidy_7.2.2 %>%
  ggplot(aes(x = reorder(pathway, NES), y = NES, fill = NES > 0)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top Enriched Hallmark Pathways: Diagnosis vs Relapse (log2FC > |1|)",
    x = "Pathway",
    y = "Normalized Enrichment Score (NES)"
  ) +
  scale_fill_manual(values = c("TRUE" = "#61BB6F", "FALSE" = "#EC7012")) +
  theme_minimal()

# Display the plot
print(plot_7.2.2)
```

### Task 7.2.3 Compare the pathways in stjudes Q4 vs DR Matss

```{r}
# Rank pathways by padj
pathways_7_2_2 <- fgsea_tidy_7.2.2$pathway
pathways_7_1_1 <- fgsea_tidy_7.1.1$pathway


# Identify common and unique pathways
common_pathways <- intersect(pathways_7_2_2, pathways_7_1_1)
unique_to_7_2_2 <- setdiff(pathways_7_2_2, pathways_7_1_1)
unique_to_7_1_1 <- setdiff(pathways_7_1_1, pathways_7_2_2)


```
#### Plotting

```{r}
# Create a list of pathways
pathway_list <- list(
  Relapse = pathways_7_2_2,
  Q4 = pathways_7_1_1
)

# Plot the Venn diagram
ggvenn(pathway_list, c("Relapse", "Q4"))
```

```{r, fig.height= 10, fig.width=10}
# Extract NES for common pathways
common_pathways_7_2_2 <- fgsea_tidy_7.2.2 %>%
  filter(pathway %in% common_pathways) %>%
  dplyr::select(pathway, NES) %>%
  mutate(dataset = "Relapse")

common_pathways_7_1_1 <- fgsea_tidy_7.1.1 %>%
  filter(pathway %in% common_pathways) %>%
  dplyr::select(pathway, NES) %>%
  mutate(dataset = "Q4")

# Combine the data
common_pathways_df <- bind_rows(common_pathways_7_2_2, common_pathways_7_1_1)

# Create the bar plot
ggplot(common_pathways_df, aes(x = reorder(pathway, NES), y = NES, fill = dataset)) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip() +
  labs(
    title = "Comparison of Common Pathways: Relapse vs Q4",
    x = "Pathway",
    y = "Normalized Enrichment Score (NES)"
  ) +
  scale_fill_manual(values = c("Relapse" = "#61BB6F", "Q4" = "#EC7012")) +
  theme_minimal()
```
#### Extracting leading genes
```{r}
# Extract leading edge genes for common pathways in Relapse dataset
leading_genes_7_2_2 <- fgsea_tidy_7.2.2 %>%
  filter(pathway %in% common_pathways) %>%
  dplyr::select(pathway, leadingEdge) %>%
  unnest(cols = c(leadingEdge)) %>%
  distinct()

# Display the leading edge genes for Relapse dataset
print(leading_genes_7_2_2)
```

#### Venn diagram for leading edfge genes

```{r}
# Function to create and save Venn diagram for a given pathway
create_and_save_venn_diagram <- function(pathway, output_dir) {
  # Extract leading edge genes for the pathway from both datasets
  leading_genes_7_2_2 <- fgsea_tidy_7.2.2 %>%
    filter(pathway == !!pathway) %>%
    dplyr::select(leadingEdge) %>%
    unnest(cols = c(leadingEdge)) %>%
    distinct() %>%
    pull(leadingEdge)
  
  leading_genes_7_1_1 <- fgsea_tidy_7.1.1 %>%
    filter(pathway == !!pathway) %>%
    dplyr::select(leadingEdge) %>%
    unnest(cols = c(leadingEdge)) %>%
    distinct() %>%
    pull(leadingEdge)
  
  # Create a list of leading edge genes
  gene_list <- list(
    Relapse = leading_genes_7_2_2,
    Q4 = leading_genes_7_1_1
  )
  
  # Plot the Venn diagram
  venn_plot <- ggvenn(
    data = gene_list,
    columns = c("Relapse", "Q4"),
    fill_color = c("#FF9999", "#99CCFF"),
    fill_alpha = 0.6,
    stroke_color = "black",
    stroke_size = 1,
    set_name_color = "black",
    set_name_size = 6,
    text_color = "black",
    text_size = 4,
    show_percentage = TRUE,
    digits = 1
  ) +
    ggtitle(paste("Venn Diagram of Leading Edge Genes for", pathway)) +
    theme(plot.title = element_text(size = 14, face = "bold"))
  
  # Save the plot to a PDF file
  ggsave(
    filename = paste0(output_dir, "/", gsub(" ", "_", pathway), ".pdf"),
    plot = venn_plot,
    width = 8,
    height = 6
  )
}

# Directory to save the plots
output_dir <- "/Users/cojulian/Desktop/Tzu_projects/Matt/Integrated_D_R_curtis/tasks_results/task7/task7.2/task7.2.3"

# Create and save Venn diagrams for each common pathway
for (pathway in common_pathways) {
  create_and_save_venn_diagram(pathway, output_dir)
}
```

#### List of genes:

```{r}

# Function to find overlapping genes for a given pathway
find_overlapping_genes <- function(pathway) {
  # Extract leading edge genes for the pathway from both datasets
  leading_genes_7_2_2 <- fgsea_tidy_7.2.2 %>%
    filter(pathway == !!pathway) %>%
    dplyr::select(leadingEdge) %>%
    unnest(cols = c(leadingEdge)) %>%
    distinct() %>%
    pull(leadingEdge)
  
  leading_genes_7_1_1 <- fgsea_tidy_7.1.1 %>%
    filter(pathway == !!pathway) %>%
    dplyr::select(leadingEdge) %>%
    unnest(cols = c(leadingEdge)) %>%
    distinct() %>%
    pull(leadingEdge)
  
  # Find overlapping genes
  overlapping_genes <- intersect(leading_genes_7_2_2, leading_genes_7_1_1)
  
  return(overlapping_genes)
}

# Create a list to store overlapping genes for each pathway
overlapping_genes_list <- list()

# Find overlapping genes for each common pathway
for (pathway in common_pathways) {
  overlapping_genes_list[[pathway]] <- find_overlapping_genes(pathway)
}

# Print the overlapping genes for each pathway
overlapping_genes_list

# Create a data frame from the list of overlapping genes
overlapping_genes_df <- bind_rows(lapply(overlapping_genes_list, function(x) {
  data.frame(Gene = x)
}), .id = "Pathway")

# Print the data frame
print(overlapping_genes_df)

write.csv(overlapping_genes_df, "/Users/cojulian/Desktop/Tzu_projects/Matt/Integrated_D_R_curtis/tasks_results/task7/task7.2/task7.2.3/overlapping_genes.csv", row.names = F)

```


## Task 7.3 Heatmap containing only chromosome 8 genes for diagnosis versus relapse of the three LH samples. Here, I want to see if GEM upregulation is just due to chromosome 8 genes generally being upregulated.

```{r}
# Connect to the Ensembl BioMart
ensembl <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")

# Get the list of genes on chromosome 8
chr8_genes <- getBM(attributes = c("hgnc_symbol", "chromosome_name"),
                    filters = "chromosome_name",
                    values = "8",
                    mart = ensembl)

# Extract the gene symbols
chr8_gene_symbols <- chr8_genes$hgnc_symbol
```

```{r}
chromosome_8_data_7.2 <- matts_df_LH_7.2 %>%
  rownames_to_column("Gene") %>% 
  filter(Gene %in% chr8_gene_symbols)


# Check for missing genes
missing_chr8_genes <- setdiff(chr8_gene_symbols, chromosome_8_data_7.2$Gene)
if (length(missing_chr8_genes) > 0) {
  cat("The following chromosome 8 genes are missing from the data:\n")
  print(missing_chr8_genes)
}



# Step 3: Separate Diagnosis and Relapse Samples
diagnosis_data_7.2 <- chromosome_8_data_7.2 %>%
  column_to_rownames("Gene") %>%  # Use Gene column as row names
  dplyr::select(matches("-D$"))

relapse_data_7.2 <- chromosome_8_data_7.2 %>%
  column_to_rownames("Gene") %>%  # Use Gene column as row names
  dplyr::select(matches("-R$"))

# Combine Diagnosis and Relapse Data
heatmap_data_7.2 <- cbind(diagnosis_data_7.2, relapse_data_7.2)
```

```{r}
# Calculate z-scores
# Calculate Z-Scores
heatmap_data_zscore_7.2 <- t(scale(t(heatmap_data_7.2)))
heatmap_data_zscore_7.2[is.na(heatmap_data_zscore_7.2)] <- 0

```


```{r}
# Step 5: Create Metadata for Heatmap Annotation
annotation <- data.frame(
  Condition = rep(c("Diagnosis", "Relapse"), each = ncol(diagnosis_data_7.2)),
  row.names = colnames(heatmap_data_zscore_7.2)
)

```

```{r, fig.width=20, fig.height=20}
pheatmap(
  heatmap_data_zscore_7.2,
  cluster_rows = F,
  cluster_cols = F,
  show_rownames = TRUE,
  show_colnames = TRUE,
  main = "Chromosome 8 Genes: Diagnosis vs Relapse (Z-score)",
  color = colorRampPalette(c("blue", "white", "red"))(50),
  annotation_col = annotation,
  fontsize_row = 13,
  fontsize = 20,
  fontsize_col = 13
)
```


### top 50 and top 100 

```{r}
# Calculate standard deviation for each gene
gene_variability <- apply(heatmap_data_zscore_7.2, 1, sd, na.rm = TRUE)

# Get top 100 genes
top_100_genes <- names(sort(gene_variability, decreasing = TRUE)[1:100])

# Get top 50 genes
top_50_genes <- names(sort(gene_variability, decreasing = TRUE)[1:50])

# Subset the data for top 100 and top 50 genes
heatmap_data_top_100 <- heatmap_data_zscore_7.2[top_100_genes, ]
heatmap_data_top_50 <- heatmap_data_zscore_7.2[top_50_genes, ]

```

```{r, fig.width=21, fig.height=20}
pheatmap(
  heatmap_data_top_100,
  cluster_rows = F,    # Enable clustering for better visualization
  cluster_cols = F,
  show_rownames = TRUE,
  show_colnames = TRUE,
  main = "Top 100 Chromosome 8 Genes: Diagnosis vs Relapse (Z-score)",
  color = colorRampPalette(c("blue", "white", "red"))(50),
  annotation_col = annotation,
  fontsize_row = 13,
  fontsize = 20,
  fontsize_col = 13
)

```


```{r, fig.width=22, fig.height=20}
pheatmap(
  heatmap_data_top_50,
  cluster_rows = F,    # Enable clustering for better visualization
  cluster_cols = F,
  show_rownames = TRUE,
  show_colnames = TRUE,
  main = "Top 50 Chromosome 8 Genes: Diagnosis vs Relapse (Z-score)",
  color = colorRampPalette(c("blue", "white", "red"))(50),
  annotation_col = annotation,
  fontsize_row = 13,
  fontsize = 20,
  fontsize_col = 13
)

```

# Task 8. Kaplan Meyer cBioPortal Pan Cancer

```{r}

```

```{r}
# Initialize connection to cBioPortal
cbio <- cBioPortal()
```

```{r}
# Get list of all studies
studies <- getStudies(cbio)

# Filter studies likely containing RNA expression data
gem_related_studies <- studies %>%
  filter(grepl("RNA", description, ignore.case = TRUE) |
         grepl("expression", description, ignore.case = TRUE))

# View relevant studies
print(gem_related_studies)

```

```{r}
# Example study ID (replace with the one of interest)
study_id <- "pog570_bcgsc_2020"  # Example

# Get molecular profiles for the study
profiles <- molecularProfiles(cbio, study_id)

pan_Multiassay = cBioDataPack(cancer_study_id = "pog570_bcgsc_2020")
```

```{r}
experiments(pan_Multiassay)
```
```{r}
colData(pan_Multiassay)[1:10, 1:10]
```

```{r}
sampleMap(pan_Multiassay)
metadata(pan_Multiassay)$name
metadata(pan_Multiassay)$description
```

```{r}
MAE = pan_Multiassay[,, "mrna_seq_rpkm"]

assay(MAE)
```


```{r, fig.height=5, fig.width=10}
pan_clinical_data = clinicalData(cbio,
studyId = study_id)

coldata = as.data.frame(colData(MAE))

# Merge the clinical data with the colData
coldata_extended <- merge(
  pan_clinical_data[, c("patientId", "AGE", "GENDER", "OS_MONTHS", "OS_STATUS")],
  coldata,
  by.x = "patientId",
  by.y = "PATIENT_ID",
  all = TRUE
)
```

### Ploting for all cancers that have more than 4 samples:

```{r}
# Directory to save the plots
output_dir <- "/Users/cojulian/Desktop/Tzu_projects/Matt/Integrated_D_R_curtis/tasks_results/task8"

# Create the directory if it does not exist
if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}

# Filter cancer types with more than 3 patients
valid_cancer_types <- names(table(pan_clinical_data$CANCER_TYPE)[table(pan_clinical_data$CANCER_TYPE) > 3])

# Loop through each valid cancer type
for (cancer_type in valid_cancer_types) {
  # Subset the clinical data for the specific cancer type
  cancer_clinical_data <- subset(pan_clinical_data, CANCER_TYPE == cancer_type)
  
  if ("GEM" %in% rownames(assays(MAE)$mrna_seq_rpkm)) {
    # Match sample IDs between assay data and clinical data
    matched_samples <- colnames(assays(MAE)$mrna_seq_rpkm) %in% cancer_clinical_data$sampleId
    
    # Extract GEM expression for matched samples
    gem_expression <- assays(MAE)$mrna_seq_rpkm["GEM", matched_samples]
    
    # Align GEM expression to clinical data
    cancer_clinical_data$GEM_expression <- gem_expression[match(
      cancer_clinical_data$sampleId,
      colnames(assays(MAE)$mrna_seq_rpkm)[matched_samples]
    )]
    
    # Categorize GEM expression into "Low" and "High" groups
    cancer_clinical_data$GEM_group <- ifelse(
      cancer_clinical_data$GEM_expression > median(cancer_clinical_data$GEM_expression, na.rm = TRUE),
      "High",
      "Low"
    )
    
    # Ensure survival status is properly coded
    cancer_clinical_data$OS_STATUS <- ifelse(grepl("LIVING", cancer_clinical_data$OS_STATUS), 0, 1)
    
    # Create the survival object
    cancer_clinical_data$y <- Surv(
      time = as.numeric(cancer_clinical_data$OS_MONTHS),
      event = cancer_clinical_data$OS_STATUS
    )
    
    # Check if there are at least 2 groups to stratify (High and Low)
    if (length(unique(cancer_clinical_data$GEM_group)) > 1) {
      # Fit Kaplan-Meier model stratified by GEM expression group
      fit <- survfit(y ~ GEM_group, data = cancer_clinical_data)
      
      # Get the total number of patients included in the plot
      num_patients <- nrow(cancer_clinical_data[!is.na(cancer_clinical_data$GEM_expression), ])
      
      # Plot the Kaplan-Meier survival curve
      g <- ggsurvplot(
        fit = fit,
        data = cancer_clinical_data,
        risk.table = TRUE,  # Add a risk table
        pval = TRUE,        # Show p-value for log-rank test
        title = paste(
          "Kaplan-Meier Survival Analysis for", cancer_type,
          "\nStratified by GEM Expression (N =", num_patients, ")"
        ),
        xlab = "Overall Survival (Months)",
        ylab = "Survival Probability",
        legend.title = "GEM Expression"
      )
      
      print(g$plot)
      
      # Save the plot as a PDF
      pdf_filename <- paste0(output_dir, "/KM_plot_", gsub(" ", "_", cancer_type), ".pdf")
      ggsave(
        filename = pdf_filename,
        plot = g$plot,
        width = 10,
        height = 6
      )
      
      cat("Plot saved for cancer type:", cancer_type, "with", num_patients, "patients\n")
    } else {
      cat("Skipping", cancer_type, "due to insufficient stratification groups.\n")
    }
  } else {
    stop("Gene 'GEM' not found in row names of the assay matrix.")
  }
}

```



### Trying bone-related cancers only

```{r}
# Filter the studies for the relevant bone cancer studies
bone_cancer_studies <- studies %>%
  filter(name %in% c(
    "Pediatric Ewing Sarcoma (DFCI, Cancer Discov 2014)",
    "Ewing Sarcoma (Institut Curie, Cancer Discov 2014)",
    "Adult Soft Tissue Sarcomas (TCGA, Cell 2017)",
    "Sarcoma (MSKCC/Broad, Nat Genet 2010)",
    "Sarcoma (MSK, Nat Commun. 2022)",
    "Leiomyosarcoma (MSK, Clin Cancer Res 2024)",
    "Soft Tissue and Bone Sarcoma (MSK, Nat Commun 2022)",
    "Sarcoma (TCGA, Firehose Legacy)",
    "Rhabdomyosarcoma (NIH, Cancer Discov 2014)",
    "Pediatric Rhabdomyosarcomas (MSK, JCO Precis Oncol 2023)",
    "Rhabdomyosarcomas (MSK, NPJ Precis Oncol 2023)"
  ))

# Print the filtered studies
print(bone_cancer_studies)

```
## Maybe look for CNA amplification of the gene in this data


## This has been a test 













